---
output: github_document
bibliography: Bibliography.bib
csl: nature.csl
header-includes: 
  - |
    \AtBeginDocument{
      \renewcommand{\r}[1]{\textcolor{red}{#1}}
      \renewcommand{\b}[1]{\textcolor{blue}{#1}}
    }
  - \usepackage{xcolor}
  - \usepackage{algorithm}
  - \usepackage{algpseudocode}
  - \usepackage{mathtools}
  - \usepackage{float}
  - \newcommand{\coleq}{\mathrel{\mathop:}=}
  - \newcommand{\ve}[1]{\textbf{#1}}
  - \newcommand{\field}[1]{\mathbb{#1}}
  - \newcommand{\defn}[1]{\emph{#1}}
  - \newcommand{\oneVect}[1]{#1}
  - \newcommand{\mat}[1]{\textbf{#1}}
  - \newcommand{\oneMat}[1]{#1}
  - \newcommand{\tensor}[1]{\mathcal{#1}}
  - \newcommand{\oneTensor}[1]{#1} 
  - \newcommand{\vs}[1]{\mathcal{#1}}
  #- \newcommand{\mn}[1]{{#1} \choose {#2}}
  #- \newcommand{\mn}[2]{\mathrel{\underset{\scriptstyle {#1}\,{#2}}{\times}}}
  #- \newcommand{\mn}[2]{\underset{\scriptsize\color{red}{#1}\,\color{blue}{#2}}{\times}}
  #- \newcommand{\mn}[2]{\mathbb{#1}}
  #- \newcommand{\mn}[2]{\sideset{_{\color{red}{#1}}}{_{\color{blue}{#2}}}{\boldsymbol{\times}}}
  - \newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}
  - \newtheorem{proposition}{Proposition}[section]
  - \newtheorem{proof}{Proof}[section]
---




```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(1)
```

# **tensortools**

Tensor operations, decompositions, and distributions in R. 

```{r}
#| message: false
library("tensortools")
library("tidyverse")
library("tictoc")
library("keras")
old_digits <- getOption("digits")
options(digits = 3)

plot_matrix <- function(input) { # function to plot matrices
  input |> 
    as_tibble(.name_repair = ~ paste0("V", seq_along(.))) |> 
    mutate(row = row_number()) |> 
    pivot_longer(-row, names_to = "col", values_to = "value") |> 
    mutate(col = as.integer(str_remove(col, "^V"))) |> 
    ggplot(aes(x = col, y = row, fill = value)) +
    geom_tile(color = "black", lwd = 0.1) +
    scale_y_reverse() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) 
}
```




# Operations



## n-mode prod

The $n$-mode matrix product of an $O$-th order tensor $\tensor{A} \in \field{k}^{I_1 \times I_2 \times \dots \times I_{n} \times \dots \times I_{O}}$ with a matrix $\mat{U} \in \field{k}^{J \times I_n}$  
is

$$(\tensor{A} \times_n \mat{U}) \in \field{k}^{I_1 \times \dots \times I_{n-1} \times J \times I_{n+1} \times \dots \times I_{O}}$$ 

with entries 

$$(\tensor{A} \times_n \mat{U})_{i_1, \dots, i_{n-1}, j, i_{n+1}, \dots i_{O}} = \sum_{i_n = 1}^{I_{N}} a_{i_{1}, i_{2}, \dots, i_{O}} u_{j, i_n}.$$ 

The tensor and matrix share one mode in common, denoted here as $I_{n}$. This operation is also called the tensor times matrix product.

The n-mode product between a tensor and a matrix can be computed with `n_mode_prod()`{.R}.

```{r}
a <- array(1:3, dim = c(3, 1, 1))
x <- matrix(4:9, nrow = 2, ncol = 3)

n_prod(a, 1, x)
```





## nm-mode prod

The $nm$-mode product between a tensor and a tensor can be computed with
`nm_prod()`{.R}. The $nm$-mode matrix product of an $O$-th order tensor $\tensor{A} \in \field{k}^{I_{1} \times I_{2} \times \dots \times I_{n} \times \dots \times I_{O}}$ with a $P$-th order tensor $\mat{U} \in \field{k}^{J_{1} \times {J_2} \times \dots \times J_{m} = I_{n} \dots \times J_{P}}$ is

$$(\tensor{A} \mn{n}{m} \tensor{U}) \in \field{k}^{I_1 \times \dots \times I_{n-1} \times I_{n+1} \times \dots \times I_{O} \times J_{1} \times \dots \times J_{m-1} \times J_{m+1} \times \dots \times J_{P}}$$ 

with entries 

$$(\tensor{A} \mn{n}{m} \tensor{U})_{i_1, \dots, i_{n-1}, k, i_{n+1}, \dots I_{O}, j_{1}, \dots, j_{m-1}, j_{m+1}, \dots j_{P}} = \sum_{i_n = 1}^{I_n} a_{i_1, i_2, \dots, i_{n}, \dots i_N} u_{j_{1}, j_{2}, \dots i_n, \dots j_{P}}.$$ 

The $n$-th mode of the first tensor matches the $m$-th mode of the second tensor, denoted here as $I_{n} = J_{m}$. This operation is also called the tensor times tensor product.

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

nm_prod(A, x, 1, 1)
```




## [Tensor product](https://en.wikipedia.org/wiki/Tensor_product)

The tensor product between a tensor and a tensor can be computed with
`tensor_prod()`{.R}. Take 2 tensors $\tensor{A} \in \field{k}^{I_{1} \times I_{2} \times \dots \times I_{O}}$ and $\tensor{U} \in \field{k}^{J_{1} \times J_{2} \times \dots \times J_{P}}$

Then the tensor product is $$(\tensor{A} \otimes \tensor{B}) \in \field{k}^{I_{1} \times I_{2} \dots \times I_{O} \times J_{1} \times \dots \times J_{P}}$$ with entries 

$$(\tensor{A} \otimes \tensor{U})_{i_1, \dots, i_{O}, j_{1}, \dots, j_{P}} = a_{i_1, i_2, \dots, i_{o}} u_{j_{1}, j_{2}, \dots \dots j_{p}}.$$ 

```{r}
A <- matrix(c(1, 2, 3, 4), nrow = 2)
x <- matrix(c(5, 6), nrow = 2)

tensor_prod(A, x)
```






# Simulating tensor variate normal draws

The density function of the multilinear normal distribution is @tensornormprop

```math
f(x) =
(2\pi)^{-p^{*}/2}
\biggl(\prod_{i=1}^{k} |\Sigma_i|^{-p^{*}/(2p_i)}\biggr)
\exp\biggl\{ -\frac{1}{2} (x-\mu)^{T} \Sigma_{1:k}^{-1} (x-\mu) \biggr\}
```

where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtnorm()`{.R} simulates random draws from the tensor variate normal with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.


```{r}
univar_draws <- rtnorm(n = 1000, mu = -2, 
                       sigmas = list(matrix(4)))

mean(simplify2array(univar_draws))

var(simplify2array(univar_draws))
```


We can also simulate random draws from a multivariate normal distribution by specifying a mean vector and the covariance matrix. The dimensions of the draws are dependent on the $\mu$ provided and the corresponding covariance matrices $\mathbf{\Sigma}$ must conform to the $\mu$ provided.

```{r}
S1 <- crossprod(matrix(data = c(1, 0.5, 0.5, 1), nrow = 2))

multivar_draws <- rtnorm(n = 1000, mu = c(2, 3), sigmas = list(S1))

multivar_draws[1:5]
```


And now we simulate from the matrix variate normal of size $2 \times 3$. By
default, the covariance matrices will be the identity.

```{r}
matrix_draws <- rtnorm(n = 1e3, mu = matrix(1:6, nrow = 2, ncol = 3))
```

When simulating draws, the output is represented a list of $n$ draws. Each element of the list represents one tensor observation. This makes it clear that the sample size indexes observations and works well with other functions in the package. Functions such as `simplify2array()` can be used to create a higher-order tensor with one mode being the sample index.


```{r}
matrix_draws[[1]]
```


Below is a simulation from a tensor variate normal of size $3 \times 4 \times
2$.The covariance structure is specified with a list of matrices. 

```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))

S1 <- crossprod(matrix(rnorm(9), nrow = 3))
S2 <- crossprod(matrix(rnorm(16), nrow = 4))
S3 <- crossprod(matrix(rnorm(4), nrow = 2))

tvn_draws <- rtnorm(n = 1e3, mu = mu_true,
                    sigmas = list(S1, S2, S3))

tvn_draws[[1]]
```



# Estimation

## MLE estimation for the tensor variate normal 

The package supports MLE estimation. Given an array of draws, it will return a list containing the MLE for the mean and covariance matrices. Earlier, we generated univariate normal draws with $\mu = 2$ and $\sigma^{2} = 4.$ Calling the function  `tensor_mle()`{.R} will return MLEs similar to the true results. 

```{r}
(univarnorm_est <- tensor_mle(draws = univar_draws, model = "normal"))
```


The multivariate normal draws had a mean vector of $\begin{pmatrix} 2 \\\\ 3 \end{pmatrix}$ and a covariance matrix $\begin{pmatrix} 1.25 & 1 \\\\ 1 & 1.25 \end{pmatrix}.$ 

```{r}
(multivarnorm_est <- tensor_mle(draws = multivar_draws, model = "normal"))
```


For the matrix variate draws, the mean matrix was $\begin{pmatrix} 1 & 3 & 5 \\\\ 2 & 4 & 6 \end{pmatrix}$ and the covariance matrices were the identity.


```{r}
matrix_est <- tensor_mle(matrix_draws, model = "normal")

matrix_est$mu
matrix_est$sigmas |> lapply(round, 3)
```


The `tensor_mle()`{.R} function also works for tensor-valued data. Below, we compare the estimated and true covariance matrices using the mean squared error (MSE).  

Also, note that there is nonidentifiability in the scaling of the covariance matrices. Given the tensor variate normal draws, we know the kronecker of all of the covariances, but not the scaling of each one. To compare the true and estimated covariance matrices, we first trace normalization. This removes arbitrary scale differences while preserving the relative covariance structure within each mode.


```{r}
tensor_est <- tensor_mle(draws = tvn_draws, model = "normal")

true_sigmas <- list(S1, S2, S3)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- tensor_est$sigmas[[i]] / sum(diag(tensor_est$sigmas[[i]]))
  
  print(mean((est_scaled - true_scaled)^2))
}
```






# Other models

To generate tensor variate skewed distributions, we will use the normal variance mean mixture model \citep{variancemeanmix} An $r$-dimensional random vector $x$ is a normal variance-mean mixture with mixing distribution $F$ if, for a given $u \ge 0$ that follows a probability distribution $F$ on $[0, \infty)$, $x|u \sim N_r(\mu + u \beta, u \Sigma)$. $\Sigma \mathbb{R}^{r \times r}$ is a constant, positive-definite matrix and $\mu \in \mathbb{R}^{r}$ and $\beta \in \mathbb{R}^{r}$ are constant vectors.

We state that $x$ is a normal variance-mean mixture with position $\mu$, drift $\beta$, structure matrix $\Sigma$, and mixing distribution $F.$ Different mixing distributions lead to different distributions. For example, if $F$, the mixing distribution, is the generalized inverse Gaussian distribution GIG($\lambda, \delta^2, \kappa^2$) then the resulting distribution of $x$ is a generalized hyperbolic distribution.

In the tensor variate case, the normal variance mean mixture model is an efficient way to introduce skewness. A random tensor $\tensor{X}$ can be written as $\tensor{X} = \tensor{M} + \mat{W} \tensor{A} + \sqrt{\mat{W}} \tensor{V}.$

## Tensor variate normal inverse Gaussian 

For the tensor variate normal inverse Gaussian distribution, we let $W \sim \operatorname{IG}(1, \kappa)$.

The density function of the normal inverse Gaussian distribution is

```math
\begin{aligned}
f_{\text{TVNIG}}(\mathcal{X}|\mathbf{V})
&= \frac{2 \exp\biggl\{\text{vec}(\mathcal{X} - \mathcal{M})^{T}
\bigotimes_{d=1}^{D} \Sigma_{d}^{-1} \text{vec}(\mathcal{A}) + \kappa \biggr\}}
{(2\pi)^{\frac{n^{*}+1}{2}} \prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}}} \\
&\quad \times
\biggl(\frac{\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1}
{\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \kappa^2} \biggr)^{-\frac{1 + n^{*}}{4}} \\
&\quad \times
K_{-\frac{1 + n^{*}}{2}} \biggl(\sqrt{\Bigl[\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \kappa^{2}\Bigr]
\Bigl[\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 1\Bigr]} \biggr).
\end{aligned}
```

where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtinvgauss()`{.R} simulates random draws from the tensor variate normal inverse Gaussian with a specified mean array mu, a list of covariance matrices, a skew array, and $\kappa$ which describes the shape of the inverse Gaussian distribution


```{r}
mu_true <- array(1:24, dim = c(3, 4, 2))
skew_true <- array(seq(0, 4, length.out = 24), dim = c(3, 4, 2))
kappa_true <- 2

invgauss_draws <- rtinvgauss(n = 1e3, mu = mu_true, skew = skew_true, 
                             sigmas = list(S1, S2, S3), kappa = kappa_true)
```


```{r}
invgauss_est <- tensor_mle(invgauss_draws, model = "invgauss")
```


For the inverse gaussian distribution, the true mean $E[X] = \tensor{M} + \tensor{A}/\kappa$. We can compare the model's estimation for the mean with the true values using MSE. 

```{r}
mean((with(invgauss_est, mu + skew/kappa) -
        (mu_true + skew_true/kappa_true))^2)
```


```{r}
mean((invgauss_est$mu - mu_true)^2)
mean((invgauss_est$skew - skew_true)^2)
invgauss_est$kappa

for(i in 1:2) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- invgauss_est$sigmas[[i]] / sum(diag(invgauss_est$sigmas[[i]]))
  
  print(mean((est_scaled - true_scaled)^2))
}
```




## Tensor variate generalized hyperbolic

For the tensor variate generalized hyperbolic distribution, we let $W \sim I(\omega, 1, \lambda)$.

The density function of the generalized hyperbolic distribution is

```math
\begin{aligned}
f_{\text{TVGH}}(\mathcal{X}|\mathbf{V})
&= \frac{\exp\biggl\{\text{vec}(\mathcal{X} - \mathcal{M})^{T}
\bigotimes_{d=1}^{D} \Sigma_{d}^{-1} \text{vec}(\mathcal{A})\biggr\}}
{(2\pi)^{\frac{n^{*}}{2}} \prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}} K_{\lambda}(\omega)} \\
&\quad \times
\biggl(\frac{\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \omega}
{\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \omega} \biggr)^{\frac{\lambda - n^{*}/2}{2}} \\
&\quad \times
K_{\lambda - n^{*}/2} \biggl(\sqrt{\Bigl[\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \omega\Bigr]
\Bigl[\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \omega\Bigr]}\biggr).
\end{aligned}
```

where $\Sigma$ is positive definite, $x \in \mathbb{R}^p$, $\mu \in \mathbb{R}^p$ and $\Sigma_{1:k} \in \mathbb{R}^p.$

The function `rtgenhyper()`{.R} simulates random draws from the tensor variate generalized hyperbolic with a specified mean array mu and a list of covariance matrices called list_sigmas.

Since the univariate normal and matrix variate normal are simpler cases of the
tensor variate normal, this function can simulate from them as well.  Here, we
simulate random draws from a univariate normal with mean $-2$ and variance $4$.

```{r}
lambda_true <- 10
omega_true <- 12
 
genhyper_draws <- rtgenhyper(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3),
                             lambda = lambda_true, omega = omega_true)
```



For the generalized hyperbolic distribution, the true mean $E[X] = \tensor{M} + \frac{K_{\lambda + 1}(\omega)}{K_{\lambda}(\omega)} \tensor{A}$. We can compare the model's estimation for the mean with the true values using MSE. 

```{r}
genhyper_est <- tensor_mle(genhyper_draws, model = "genhyper", quiet = FALSE, tol = 1e-5)

mean((
  with(genhyper_est, mu + besselK(x = omega, nu = lambda + 1)/
                    besselK(x = omega, nu = lambda) * skew) -
    (mu_true + besselK(x = omega_true, nu = lambda_true + 1)/
       besselK(x = omega_true, nu = lambda_true) * skew_true)
  )^2)

mean((genhyper_est$mu - mu_true)^2)
mean((genhyper_est$skew - skew_true)^2)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- genhyper_est$sigmas[[i]] / sum(diag(genhyper_est$sigmas[[i]]))
  
  print(mean((est_scaled - true_scaled)^2))
}

genhyper_est$lambda
genhyper_est$omega
```


## Tensor variate variance gamma

For the tensor variate variance gamma distribution, we let $W \sim \text{Gamma}(\gamma, \gamma)$.

The density function of the variance gamma distribution is

```math
\begin{aligned}
f_{\text{TVVG}}(\mathcal{X}|\mathbf{V})
&= \frac{2\gamma^{\gamma} \exp\biggl\{\text{vec}(\mathcal{X} - \mathcal{M})^{T}
\bigotimes_{d=1}^{D} \Sigma_{d}^{-1} \text{vec}(\mathcal{A})\biggr\}}
{(2\pi)^{\frac{n^{*}}{2}} \prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}} \Gamma(\gamma)} \\
&\quad \times
\biggl(\frac{\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1})}
{\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 2\gamma} \biggr)^{\frac{\gamma - n^{*}/2}{2}} \\
&\quad \times
K_{\gamma - n^{*}/2} \biggl(\sqrt{\Bigl[\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + 2\gamma\Bigr]
\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1})} \biggr).
\end{aligned}
```

```{r}
vargamma_draws <- rtvargamma(n = 1e3, mu = mu_true, skew = skew_true,
                             sigmas = list(S1, S2, S3))
```


```{r}
vargamma_est <- tensor_mle(vargamma_draws, model = "vargamma", quiet = FALSE)

mean((with(vargamma_est, mu + skew) - (mu_true + skew_true))^2)

mean((vargamma_est$mu - mu_true)^2)
mean((vargamma_est$skew - skew_true)^2)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- vargamma_est$sigmas[[i]] / sum(diag(vargamma_est$sigmas[[i]]))
  
  print(mean((est_scaled - true_scaled)^2))
}

vargamma_est$gamma
```

## Tensor variate skewed t

For the tensor variate skewed t distribution, we let $W \sim \operatorname{Inv\text{-}Gamma}(\nu/2, \nu/2)$.

The density function of the skewed t distribution is

```math
\begin{aligned}
f_{\text{TVST}}(\mathcal{X}|\mathbf{V})
&= \frac{2\bigl(\frac{\nu}{2}\bigr)^{\nu/2} \exp\biggl\{\text{vec}(\mathcal{X} - \mathcal{M})^{T}
\bigotimes_{d=1}^{D} \Sigma_{d}^{-1} \text{vec}(\mathcal{A})\biggr\}}
{(2\pi)^{\frac{n^{*}}{2}} \prod_{d=1}^{D} |\Sigma_{d}|^{\frac{n^{*}}{2n_{d}}} \Gamma\bigl(\frac{\nu}{2}\bigr)} \\
&\quad \times
\biggl(\frac{\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \nu}
{\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1})} \biggr)^{-\frac{\nu + n^{*}}{4}} \\
&\quad \times
K_{-\frac{\nu + n^{*}}{2}} \biggl(\sqrt{\rho(\mathcal{A}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1})
\Bigl[\delta(\mathcal{X}; \mathcal{M}, \bigotimes_{d=1}^{D} \Sigma_{d}^{-1}) + \nu\Bigr]} \biggr).
\end{aligned}
```

```{r}
nu_true <- 20
skew_true <- array(c(
   0.6,  0.3, -0.2,  0.5,
   0.4, -0.5,  0.2, -0.3,
  -0.4,  0.7,  0.1, -0.6,

   0.5, -0.2,  0.3,  0.4,
  -0.3,  0.6, -0.4,  0.2,
   0.1, -0.5,  0.7, -0.1
), dim = c(3, 4, 2))

skewt_draws <- rtskewt(n = 1e3, mu = mu_true, skew = skew_true,
                       sigmas = list(S1, S2, S3), nu = nu_true)
```


```{r}
skewt_est <- tensor_mle(skewt_draws, model = "skewt",
                        quiet = FALSE)

mean((with(skewt_est, mu + (nu)/(nu-2) * skew) -
        (mu_true + nu_true / (nu_true - 2) * skew_true))^2)

for(i in 1:3) {
  true_scaled <- true_sigmas[[i]] / sum(diag(true_sigmas[[i]]))
  est_scaled <- skewt_est$sigmas[[i]] / sum(diag(skewt_est$sigmas[[i]]))
  
  print(mean((est_scaled - true_scaled)^2))
}
```




## Assessing tensor variate normality

We assess tensor variate normality using the Mahalanobis distance. If $\mathcal{X}$ follows an order-$O$ tensor variate normal distribution, then $\text{vec}(\mathcal{X})$ follows a multivariate normal distribution with mean $\text{vec}(\mathcal{M})$ and covariance matrix $\boldsymbol{\Sigma}_{O} \otimes \cdots \otimes \boldsymbol{\Sigma}_{1}$. This is why we can vectorize the tensor draws and compare the multivariate Mahalanobis distances to the tensor variate Mahalanobis distances. If we have properly reconstructed the structure using the MLEs $\hat{\mathcal{M}}$ and $\hat{\Sigma}_{O}, \cdots \hat{\Sigma}_{1}$, then the standardized draws should be from a standard normal. 


If a tensor normal covariance structure is present, then the multivariate and tensor variate distances should be close to one another. Following the matrix variate normality paper, this can be visualized with a distance-distance plot: when the tensor normal structure is appropriate, the distances lie close to the reference line; when the Kronecker product structure is absent, the distances diverge.


The function `mahalanobis_dist()`{.R} computes both distances using fitted MLEs, and `mahalanobis_test()`{.R} performs a two-sample Kolmogorov-Smirnov test to compare their empirical distributions.

```{r}
A <- matrix(rnorm(6^2), 6, 6)
s_vec <- crossprod(A)   # positive definite, general covariance

vec_draws <- rtnorm(n = 1e3, mu = 1:6, sigmas = list(s_vec))

matrix_nocovar_draws <- lapply(1:1e3, function(i) {
  array(vec_draws[[i]], dim = c(2, 3))
})
```


```{r}
plot_malanobis <- function(draws, title = "Dist") {
  distances <- mahalanobis_dist(draws)
  test <- mahalanobis_test(distances)
  plot_dist(distances)
  title(main = title)
  with(test, mtext(text = sprintf("D = %.3f, p = %.3f", statistic, p.value)), 
       side = 3, line = 1)
}
```


```{r}
par(mfcol = c(1, 4))
plot_malanobis(multivar_draws, title = "Multivar")
plot_malanobis(matrix_draws, title = "Mat norm")
plot_malanobis(matrix_nocovar_draws, title = "Mat no covar norm")
plot_malanobis(tvn_draws, title = "Tensor norm")
```

We are also interested in performing a likelihood ratio test. We are curious if one of the covariance matrices might be the identity matrix. Thus, we run the algorithm and restrict these matrices. 

For a restriction set \(\mathcal{R} \subset \{1,\ldots,D\}\), the restricted model fixes \(\Sigma_r = I_{n_r}\) for every \(r \in \mathcal{R}\), while all other covariance matrices are estimated. The likelihood ratio test compares 

$$H_0: \Sigma_r = I_{n_r} \ \forall r \in \mathcal{R} \quad \text{vs.} \quad H_A: \Sigma_r \neq I_{n_r} \ \text{for at least one } r \in \mathcal{R}.$$

The test statistic is 

$$2\{\ell(\hat\theta_{\mathrm{full}}) - \ell(\hat\theta_{\mathcal{R}})\}
\sim \chi^2_{\nu_{\mathcal{R}}}, \quad \nu_{\mathcal{R}} = \sum_{r \in \mathcal{R}} \biggl\{\frac{n_r(n_r+1)}{2} - 1\biggr\}$$

If the $p$-value is large, then we fail to reject the null hypothesis and conclude the restricted model fits the data as well as the unrestricted model. If the $p$-value is small, we reject the null hypothesis, which states the unrestricted model has a significantly better fit than the restricted model.

```{r}
true_str <- kronecker(S3, S2) |> kronecker(S1)
  
test_restrict <- function(draws, restrict) {
  curr_restrict <- tensor_mle(draws, restrict = restrict, model = "normal")
  
  if(is.null(restrict)) restrict <- "None"

  curr_str <- with(curr_restrict, kronecker(sigmas[[3]], sigmas[[2]]) |>
                                  kronecker(sigmas[[1]]))

  tibble(mse = mean((curr_str - true_str)^2),
         loglik = curr_restrict$loglik,
         k = curr_restrict$k,
         BIC = curr_restrict$BIC,
         restrict = paste(restrict, collapse = ", "))
}

(res <- map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = tvn_draws, test_restrict))
```


```{r}
lrt_test <- function(draws, restrict) {
  full_model <- tensor_mle(draws, model = "normal")
  restrict_model <- tensor_mle(draws, model = "normal", restrict = restrict)
  
  pval <- pchisq(2 * (full_model$loglik - restrict_model$loglik), 
                 df = full_model$k - restrict_model$k, lower.tail = FALSE)
  
  tibble(restrict = paste(restrict, collapse = ", "), 
         pval = pval)
}

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = tvn_draws, lrt_test)
```


```{r}
true_str <- kronecker(S3, diag(4)) |> kronecker(S1)

second_identity <- rtnorm(n = 1e3, mu = mu_true, 
                          sigmas = list(S1, diag(4), S3))

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)), 
        draws = second_identity, test_restrict)

map_dfr(list(NULL, c(1), c(2), c(3), c(1, 2), c(1, 3), c(2, 3)),
        draws = second_identity, lrt_test)
```





## Tensor Variate Skewed T


```{r}
tibble(skew25 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(0.25, nrow = 2, ncol = 3), nu = 6)),
       skew50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(0.5, nrow = 2, ncol = 3), nu = 6)),
       skew0 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(0, nrow = 2, ncol = 3), nu = 6)),
       skew1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(1, nrow = 2, ncol = 3), nu = 6)),
       skew5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                              skew = matrix(5, nrow = 2, ncol = 3), nu = 6)),
       skew10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                               skew = matrix(10, nrow = 2, ncol = 3), nu = 6))) |>
  pivot_longer(cols = everything()) |> 
  mutate(name = factor(name, levels = c("skew0", "skew1", 
                                     "skew5", "skew10", "skew25", "skew50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()

tibble(nu05 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 0.5)),
       nu1 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                            skew = matrix(1, nrow = 2, ncol = 3), nu = 1)),
       nu5 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                            skew = matrix(1, nrow = 2, ncol = 3), nu = 5)),
       nu10 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 10)),
       nu20 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 20)),
       nu50 = unlist(rtskewt(n = 1e3, mu = matrix(0, nrow = 2, ncol = 3), 
                             skew = matrix(1, nrow = 2, ncol = 3), nu = 50))) |>
  pivot_longer(cols = everything()) |>
  mutate(name = factor(name, levels = c("nu05", "nu1", "nu5", "nu10", "nu20", "nu50"))) |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "pink") +
  facet_wrap(~name, nrow = 2, scales = "free_x") +
  theme_minimal()
```

Suppose \(X_{n \times p} \sim N_{np}(0, \Sigma_1, \Sigma_2)\). A natural question is whether the usual sample covariance of the rows can estimate \(\Sigma_2\), and whether the sample covariance of the columns can estimate \(\Sigma_1\).

If we compute the usual row covariance

$$S_{\text{row}} = \frac{1}{n-1} X^T \biggl(I_n - \frac{1}{n}\mathbf{1}_n \mathbf{1}_n^T\biggr) X,$$

then under the matrix normal model,

$$\mathbb{E}(S_{\text{row}})
= \frac{\operatorname{tr}\left[\left(I_n - \frac{1}{n}\mathbf{1}_n \mathbf{1}_n^T\right)\Sigma_1\right]}{n-1}\Sigma_2.$$

So the usual sample covariance does recover the \emph{shape} of \(\Sigma_2\), but only up to a multiplicative constant determined by \(\Sigma_1\). The same argument with \(X^T\) shows that the usual column covariance recovers the shape of \(\Sigma_1\), but only up to a multiplicative constant determined by \(\Sigma_2\).

The simulation below shows exactly that. The trace-normalized naive covariance matrices are close to the truth in all three cases, but the raw traces are distorted by the covariance matrix on the other mode.

```{r}
mse <- function(mat1, mat2) {
  mat1_scale <- mat1/diag(mat1)
  mat2_scale <- mat2/diag(mat2)
  
  mean((mat1_scale - mat2_scale)^2)
}

compare_naive_covs <- function(sigmas, n_rep = 200,
                               n_row = 30, n_col = 4) {
  
  draws <- rtnorm(n = n_rep, mu = matrix(0, nrow = n_row, ncol = n_col),
                  sigmas = sigmas)

  draws_mat <- matrix(aperm(simplify2array(draws), c(1, 3, 2)), ncol = 4)
  transpose_draws_mat <- matrix(aperm(simplify2array(draws), c(2, 1, 3)), ncol = 30)
  
  s2_naive <- var(draws_mat)
  s1_naive <- var(transpose_draws_mat)
  
  mle_est <- tensor_mle(draws, model = "normal")
  
  s2_hat <- mle_est$sigmas[[2]]
  s1_hat <- mle_est$sigmas[[1]]

  tibble(name = c("s2_naive", "s1_naive", "s2_hat", "s1_hat"),
         error = c(mse(s2_naive, sigmas[[2]]),
                   mse(s1_naive, sigmas[[1]]),
                   mse(s2_hat, sigmas[[2]]),
                   mse(s1_hat, sigmas[[1]])),
         trace_ratio = c(sum(diag(s2_naive)) / sum(diag(sigmas[[2]])),
                         sum(diag(s1_naive)) / sum(diag(sigmas[[1]])),
                         sum(diag(s2_hat)) / sum(diag(sigmas[[2]])),
                         sum(diag(s1_hat)) / sum(diag(sigmas[[1]]))))
}

set.seed(123)

s1 <- crossprod(matrix(rnorm(30^2), nrow = 30))
s2 <- crossprod(matrix(rnorm(4^2), nrow = 4))

rbind(
  compare_naive_covs(list(diag(30), s2)) |> mutate(setting = "Sigma1 = I"),
  compare_naive_covs(list(s1, diag(4))) |> mutate(setting = "Sigma2 = I"),
  compare_naive_covs(list(s1, s2)) |> mutate(setting = "both non-identity")
) |> View()
```

In this table, `s2_error` measures how close `var(X)` is to the shape of \(\Sigma_2\) after trace normalization, and `s1_error` does the same for `var(t(X))` and \(\Sigma_1\). The trace ratios show the scale distortion. For example, when \(\Sigma_2 = I\), the usual covariance of the rows still recovers the shape of \(\Sigma_2\) well, but its trace is inflated by a factor of about \(29\), coming from \(\Sigma_1\).

- `var(X)` can estimate the shape of \(\Sigma_2\), but not its absolute scale, unless the scale contribution from \(\Sigma_1\) is fixed.
- `var(t(X))` can estimate the shape of \(\Sigma_1\), but not its absolute scale, unless the scale contribution from \(\Sigma_2\) is fixed.
- This is exactly why the tensor and matrix normal models need an identifying constraint such as fixing the trace of one mode covariance.






```{r}
make_skew <- function(dims, strength = 1) {
  A <- array(rnorm(prod(dims)), dim = dims)
  A / sqrt(sum(A^2)) * strength
}
```


```{r}
s1 <- crossprod(matrix(rnorm(4), nrow = 2))
s2 <- crossprod(matrix(rnorm(9), nrow = 3))
s3 <- crossprod(matrix(rnorm(16), nrow = 4))

skew_true <- make_skew(dims = c(2, 3, 4), strength = 1.5)

draws <- rtskewt(n = 1e3, mu = array(1:24, dim = c(2, 3, 4)), 
                 sigmas = list(s1, s2, s3), skew = skew_true,
                 nu = 40)

model_compare <- function(draws) {
  res_normal <- tensor_mle(draws, model = "normal", quiet = FALSE)
  res_skewt <- tensor_mle(draws, model = "skewt", quiet = FALSE)
  res_vargam <- tensor_mle(draws, model = "vargamma", quiet = FALSE)
  res_invgauss <- tensor_mle(draws, model = "invgauss", quiet = FALSE)
  res_genhyper <- tensor_mle(draws, model = "genhyper", quiet = FALSE)

  all_mod <- 
    tibble(model = c("Normal", "Skewt", "Vargamma", "Invgauss", "Genhyper"),
    loglik = c(res_normal$loglik, res_skewt$loglik,
               res_vargam$loglik, res_invgauss$loglik, res_genhyper$loglik),
    k = c(res_normal$k, res_skewt$k,
          res_vargam$k, res_invgauss$k, res_genhyper$k),
    AIC = c(res_normal$AIC, res_skewt$AIC,
            res_vargam$AIC, res_invgauss$AIC, res_genhyper$AIC),
    BIC = c(res_normal$BIC, res_skewt$BIC,
            res_vargam$BIC, res_invgauss$BIC, res_genhyper$BIC), 
    mod_obj = list(res_normal, res_skewt, res_vargam, res_invgauss, res_genhyper)
  )
  
  all_mod
}
```



## Example: CIFAR Data


Now that we have set up our tensor distributions, we can use them on a real-world example. The CIFAR 100 dataset contains images of a set of 100 common objects. In this case, we will only be using the maple trees, which have an ID value of 48.
```{r}
library(torchvision)
library(patchwork)

plot_image <- function(img) {
  rgb_image <-
    expand_grid(x = seq_len(dim(img)[1]),
                y = seq_len(dim(img)[2])) |>
    mutate(r = as.vector(img[, , 1]), 
           g = as.vector(img[, , 2]),
           b = as.vector(img[, , 3]),
           fill = rgb(r, g, b))
  
  ggplot(rgb_image, aes(x = x, y = y, fill = fill)) +
    geom_raster(interpolate = TRUE) +
    scale_fill_identity() +
    coord_fixed(expand = FALSE) +
    scale_y_reverse() +
    theme_void()
}
```


Once we have the CIFAR dataset loaded in from torchvision, we can plot 20 example images. The images are stored as an array of size 32 x 32 x 3. This is an image that is 32 x 32 pixels stored as red, green, and blue values. 
```{r}
cifar100 <- cifar100_dataset(download = TRUE)

maple_idx <- which(cifar100$y == 48)

maple_trees <- vector("list", length(maple_idx))

# store as a list of images
for (i in seq_along(maple_idx)) {
  maple_trees[[i]] <- cifar100[[maple_idx[i]]]$x/255
}

wrap_plots(lapply(maple_trees[1:20], plot_image), ncol = 5)
```


We can fit the best model to this set of images. The function `model_compare()`{.R} takes in a dataset and fits all of the possible tensor models. It returns the result as a data frame with columns containing the name of the model, log likelihood, number of parameters denoted as k, AIC, BIC, and the model object itself. We can then select the model with the lowest BIC to proceed. 
```{r}
res <- model_compare(maple_trees)

print(res)

best_mod <- res$mod_obj[[which.min(res$BIC)]]
```


In this case, the best model was the normal inverse Gaussian. The expected value of its draws is $\mathcal{M} + \mathcal{A}/\kappa$, where $\mathcal{M}$ is the location parameter and $\mathcal{A}$ is the skew parameter. The plot below contains three images: the first is the empirical average of all of the maple trees without a model assumption. The second is the location parameter $\mathcal{M}$ from the model. Lastly, we have $E[X]$, which is the expected average from the model. 

$\mathcal{M}$ is closer to a traditional green maple tree. This seems to be a good starting point to model the variety of colors of the maples, from green to red. Once we add the skew parameter, the $E[X]$ appears similar to the empirical average.. 
```{r}
plot_image(apply(simplify2array(maple_trees), 1:3, mean)) + 
  labs(title = "Avg")+
plot_image(best_mod$mu) +
  labs(title = "M") +
plot_image(with(best_mod, mu + skew/kappa)) +
  labs(title = "E[X]")
```


We can also create summary plots of the skew $\mathcal{A}$ and the covariance matrices $\Sigma$ from the model. We converted the covariance matrices to correlation matrices to make it easier to interpret each matrix, since this places them on the same scale. 
```{r}
plot_matrix(best_mod$skew[, , 1]) +
  plot_matrix(best_mod$skew[, , 2]) +
  plot_matrix(best_mod$skew[, , 3]) + 
  plot_layout(guides = 'collect') &
  scale_fill_gradient2(#low = "blue", mid = "white", high = "red",
    midpoint = 0,
    limits = c(min(best_mod$skew), max(best_mod$skew)))

sig_corr <- lapply(best_mod$sigmas, cov2cor)

plot_matrix(sig_corr[[1]]) +
  plot_matrix(sig_corr[[2]]) +
  plot_matrix(sig_corr[[3]]) +
  plot_layout(guides = "collect") &
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1))
```



Now that we have the best model for the maple trees, we can obtain the likelihood of each image. 
```{r}
all_lik <- vector(mode = "numeric", length = 500)

for(i in 1:500) {
  all_lik[i] <- dtinvgauss(x = maple_trees[[i]], mu = best_mod$mu,
                           skew = best_mod$skew, sigmas = best_mod$sigmas, 
                           kappa = best_mod$kappa, log = TRUE)
}

all_lik |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value), color = "black", fill = "lightpink",
                 bins = 40) +
  theme_minimal() +
  labs(x = "Likelihood", y = "Count")
```


There are clearly some outliers, both at much smaller and much larger likelihoods. Out of curiosity, I plotted the lowest likelihoods. All of these maples are predominantly orange and red, which differ from the average tensor that the model had. It seems that the data contains many more green maples, so the model struggles to account for the possible variation in tree color. 
```{r}
wrap_plots(lapply(maple_trees[which(all_lik <= 4500)], plot_image), 
           ncol = 5)
```


As expected, all of the maples that had a large likelihood follow the pattern of being a forest green and mainly in the center of the image. 
```{r}
wrap_plots(lapply(maple_trees[which(all_lik >= 9000)], plot_image), 
           ncol = 5)
```


```{r}
# classes <- sort(unique(y))
# 
# best_models <- vector(mode = "list", length = 10L)
# 
# for(class in classes) {
#   class_x <- x[y == class]
#   
#   curr_model <- model_compare(class_x[1:300])
#   
#   best_mod <- curr_model$mod_obj[[which.min(curr_model$BIC)]]
#   
#   best_models[[class]] <- best_mod
# }
# 
# View(best_models)
```


```{r}
# cifar_indices <- which(cifar100$y %in% 40:48)
# 
# n <- length(cifar_indices)
# 
# imgs <- vector("list", n)
# 
# for (i in seq_len(n)) {
#   imgs[[i]] <- cifar100[[cifar_indices[i]]]$x/255
# }
# 
# class <- cifar100$y[cifar_indices]



#imgs <- imgs[1:100]

```


## Installation

You can install the development version of tensortools from
[GitHub](https://github.com/) with:
    
```{r}
#| eval: false
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("ankan12/tensortools")
```


Reset digits

```{r}
options(digits = old_digits)
```
