---
title: "Rosenblatt Derivation"
format: html
---

\newcommand{\coleq}{\mathrel{\mathop:}=}
\newcommand{\ve}[1]{\textbf{#1}}
\newcommand{\field}[1]{\mathbb{#1}}
\newcommand{\defn}[1]{\emph{#1}}
\newcommand{\oneVect}[1]{#1}
\newcommand{\mat}[1]{\textbf{#1}}
\newcommand{\oneMat}[1]{#1}
\newcommand{\tensor}[1]{\mathcal{#1}}
\newcommand{\oneTensor}[1]{#1}
\newcommand{\vs}[1]{\mathcal{#1}}
\newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}

```{r, message = FALSE, echo = FALSE}
library(tidyverse)
library(tensortools)
library(invgamma)
library(patchwork)
```

## Rosenblatt transformation

The Rosenblatt transformation is the multivariate extension of the probability integral transformation. It applies the conditional CDF of each element given the elements before it. This removes the modeled dependence sequentially. If the proposed joint distribution is correct, the resulting transformed data are independent Uniform(0, 1) random variables.

Let

$$
X=(X_1,\ldots,X_p)^{T}
$$

be a continuous random vector with joint distribution function $F(x_1, \dots, x_k)$. 

We define the Rosenblatt transformation as follows:

$$
\begin{aligned}
U_1 &=F_{X_1}(X_1),\\
U_2 &=F_{X_2\mid X_1}(X_2\mid X_1),\\
U_3 &=F_{X_3\mid X_1,X_2}(X_3\mid X_1,X_2),\\
&\ \vdots\\
U_p &=F_{X_p\mid X_1,\ldots,X_{p-1}} (X_p\mid X_1,\ldots,X_{p-1}).
\end{aligned}
$$

If the assumed joint distribution is correct and its parameters are known,
then

$$
U_1,\ldots,U_p \stackrel{\mathrm{ind}}{\sim} \text{Uniform}(0,1).
$$

Each random variable is transformed using its distribution conditional on all the random variables that came before it.

## Skewed tensor variate distributions 

The tensor skew-$t$, variance-gamma, inverse-Gaussian, and generalized
hyperbolic distributions implemented in `tensortools` share the representation

$$
\mathcal X = \mathcal M + W\mathcal A+ \sqrt{W}\,\mathcal Z,
$$

where $\mathcal A$ is the skew tensor, $W$ is a positive scalar mixing
variable, and

$$
\mathcal Z\sim
\operatorname{TVN}(\mathbf 0,\Sigma_1,\ldots,\Sigma_D)
$$

is independent of $W$. Conditional on $W=w$,

$$X \sim N(\mathcal{M} + w \mathcal{A}, w [\Sigma_{D} \otimes \dots \Sigma_{1}])$$

## Univariate Skewed-t 

Take $X = \mu + W A + \sigma \sqrt{W} Z, \quad Z \sim N(0, 1)$ where $W > 0$ is a latent mixing variable. Conditional on $W = w, X|W = w \sim N(\mu + wa, w \sigma^2)$. That is, if we knew the value $w$ of the latent mixing variable, $X$ would follow a normal distribution with mean $\mu + w a$ and variance $w \sigma^{2}$. 

In the univariate case, the Rosenblatt transformation is the same as the probability integral transform. 

First, we can obtain the marginal CDF of $X$ by integrating out the dependence on $W$.  
$$
\begin{aligned}
F_{X}(x) &= P(X \le x) \\ 
&= \int_{0}^{\infty} P(X \le x|W = w) f_{W}(w) dw \\
&= \int_{0}^{\infty} f_{X|W}(x|w) f_W(w) dw \\
&= \int_{0}^{\infty} \Phi\left(\frac{x-\mu-wa}{\sigma \sqrt{w}}\right) f_W(w) dw
\end{aligned}
$$ 


Then, we can apply this CDF to the draws. If the model is correct, $U \sim \text{Uniform}(0, 1)$. 

```{r}
nu <- 4 # generate univariate values
mu <- 1
variance <- 2
skew <- 3

# create univariate draws
univar_skewt <- rtskewt(n = 1e4, mu = mu, sigmas = list(matrix(variance)), 
                        skew = skew, nu = nu)

skew_t_cdf <- function(x, mu, variance, skew, nu) {
  shape <- nu/2
  rate <- nu/2

  integrand <- function(w) {
    pnorm((x - mu - w * skew) / sqrt(w * variance)) *  # f(x|w)
    dinvgamma(w, shape = shape, rate = rate) # f(w)
  }

  # integrate out dependence on w
  integrate(integrand, lower = 0, upper = Inf)$value
}

res <- numeric(length(univar_skewt))

for (i in seq_along(univar_skewt)) { # apply to each draw
  res[i] <- skew_t_cdf(x = univar_skewt[i], mu = mu, variance = variance,
                       skew = skew, nu = nu)
}

ggplot() +
  geom_histogram(aes(x = res, y = after_stat(density)), 
                 fill = "white", color = "black", 
                 breaks = seq(0, 1, length.out = 51)) +
  geom_function(fun = dunif, color = "red", lty = 2) +
  labs(x = "Value", y = "Density") +
  theme_minimal() 
```


## Multivariate Skewed-t

Now suppose we observe a bivariate skewed-t.

$X = \mu + W \mathbf{a} + \sqrt{W} \mathbf{Z}, \quad Z \sim N_{2}(0, \Sigma)$

where 

$$\Sigma = 
\begin{pmatrix}
\sigma_{1}^{2} & \rho \sigma_{1} \sigma_{2} \\
\rho \sigma_{1} \sigma_{2} & \sigma_{2}^{2}
\end{pmatrix}
$$

Now we need the Rosenblatt to compute $U_{1} = F_{X_{1}}(x_{1})$ and $U_{2} = F_{X_{2}|X_{1}}(x_{2}|x_{1}).$

### Multivariate Skewed-t $X_{1}$ 

First, consider $X_{1}$. Conditional on $W = w, X_{1}|W = w \sim N(\mu_{1} + w a_{1}, w \sigma_{1}^{2})$. This is the same as the univariate case we defined above. 

```{r}
nu <- 4 # generate bivariate values
mu <- c(1, 2)
variance <- matrix(data = c(1, 0.5, 0.5, 3), nrow = 2)
skew <- c(0.2, 0.4)

bivar_skewt <- rtskewt(n = 1e4, mu = mu, sigmas = list(variance),
                       skew = skew, nu = nu)
```

### Multivariate Skewed-t $X_{2}$ 

Once we have observed $X_{1} = x_{1}$, we have more information about the mixing variable $W$. Instead of using $f_{W}(w)$, we want to find $f_{W|X_{1}}(w|x_{1})$. 

We can use Bayes' rule to compute 

$$f_{W|X_{1}}(w|x_{1}) \propto f_{X_{1}|W}(x_{1}|w) f_{W}(w)$$

We can then compute $U_{2}$ using the updated CDF of $X_{2}$. 

$$F_{X_{2}|X_{1}}(x_{2}|x_{1}) = \int_{0}^{\infty} F_{X_{2}|X_{1}, W)}(x_{2}|x_{1}, w) f_{W|X_{1}}(w|x_{1}) dw$$

We can apply the CDF to the draws. If the model is correct, they should become Uniform(0, 1). 

```{r}
n <- n_draws(bivar_skewt)

L <- t(chol(variance))

b <- drop(forwardsolve(L, skew))

centered <- bivar_skewt - mu

Y <- t(forwardsolve(L, t(centered)))

conditional_u2 <- function(x1, x2, b1, b2, nu) {
  weight <- function(w) {
    dnorm(x1, mean = w * b1, sd = sqrt(w)) *
      dinvgamma(w, shape = nu / 2, rate = nu / 2)
  }

  numerator <- integrate(
    function(w) {
      pnorm(x2, mean = w * b2, sd = sqrt(w)) * weight(w)
    },
    lower = 0,
    upper = Inf
  )$value

  denominator <- integrate(
    weight,
    lower = 0,
    upper = Inf
  )$value

  numerator / denominator
}

u1 <- vapply(
  seq_len(n), 
  function(i) {
    skew_t_cdf(
      x        = bivar_skewt[i][1],
      mu       = mu[1],
      variance = variance[1, 1],
      skew     = skew[1],
      nu       = nu
    )
  },
  numeric(1)
)

# integrate() itself is not vectorized over observations
u2 <- vapply(
  seq_len(n),
  function(i) {
    conditional_u2(
      x1 = Y[i, 1],
      x2 = Y[i, 2],
      b1 = b[1],
      b2 = b[2],
      nu = nu
    )
  },
  numeric(1)
)

res_whitened <- cbind(u1 = u1, u2 = u2)
```

```{r}
res_whitened |> 
  as.numeric() |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value, y = after_stat(density)), fill = "white", color = "black", breaks = seq(0, 1, length.out = 51)) +
  geom_function(fun = dunif, color = "red", lty = 2) +
  labs(x = "Value", y = "Density") +
  theme_minimal()
```


```{r}
draws_orig <- tibble(x1 = rep(0, 1e4), x2 = rep(0, 1e4))

for(i in 1:1e4) {
  draws_orig$x1[i] <- bivar_skewt[i][1]
  draws_orig$x2[i] <- bivar_skewt[i][2]
}

draws_orig |> 
  ggplot() +
  geom_point(aes(x = x1, y = x2)) +
  theme_minimal() +
  labs(x = "X1", y = "X2") +
draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1, y = x2)) +
  theme_minimal() +
  labs(x = "U1", y = "X2") +
draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1, y = u2)) +
  theme_minimal() +
  labs(x = "U1", y = "U2")
```


## Wrong Parameters

```{r}
#true variance: c(1, 0.5, 0.5, 3)
#replace true parameter value
variance <- matrix(data = c(3, 0.5, 0.5, 3), nrow = 2)

L <- t(chol(variance))

b <- drop(forwardsolve(L, skew))

centered <- bivar_skewt - mu

Y <- t(forwardsolve(L, t(centered)))

u1 <- vapply(
  seq_len(n), 
  function(i) {
    skew_t_cdf(
      x        = bivar_skewt[i][1],
      mu       = mu[1],
      variance = variance[1, 1],
      skew     = skew[1],
      nu       = nu
    )
  },
  numeric(1)
)

# integrate() itself is not vectorized over observations
u2 <- vapply(
  seq_len(n),
  function(i) {
    conditional_u2(
      x1 = Y[i, 1],
      x2 = Y[i, 2],
      b1 = b[1],
      b2 = b[2],
      nu = nu
    )
  },
  numeric(1)
)

res_whitened_wrong <- cbind(u1 = u1, u2 = u2)

res_whitened_wrong |> 
  as.numeric() |> 
  as_tibble() |> 
  ggplot() +
  geom_histogram(aes(x = value, y = after_stat(density)), fill = "white", color = "black", 
                     breaks = seq(0, 1, length.out = 51)) +
  geom_function(fun = dunif, color = "red", lty = 2) +
  labs(x = "Value", y = "Density") +
  theme_minimal() 
```


We need to assess whether the transformed draws are uniform on the unit hypercube.  

# Test 2017YMQ Method

```{r}
library(SHT)

U <- as.matrix(res_whitened)

unif.2017YMq(U)

U_wrong <- as.matrix(res_whitened_wrong)

unif.2017YMq(U_wrong)
```

# Test Coverage as a Metric

```{r}
library(DiceDesign)

U <- U_matrix

stopifnot(
  is.matrix(U),
  all(is.finite(U)),
  all(U >= 0),
  all(U <= 1)
)
```

```{r}
tic()
discrepancy_result <- discrepancyCriteria(
  U_wrong,
  type = c("L2star")
)
toc()

coverage(U)

coverage(U_wrong)
#0.0034
#0.0459
```

```{r}
C_obs <- coverage(U_wrong)

tic()
C_null <- replicate(B, {
  U0 <- matrix(runif(nrow(U_wrong) * ncol(U_wrong)),
               nrow = nrow(U_wrong))
  coverage(U0)
})
toc()

p_upper <- (1 + sum(C_null >= C_obs)) / (B + 1)

d <- ncol(U)

coverage_expected <- sqrt(
  gamma(1 + 2/d) / gamma(1 + 1/d)^2 - 1
)

coverage_fast <- function(X) {
  X <- as.matrix(X)

  nn <- RANN::nn2(
    data = X,
    query = X,
    k = 2
  )$nn.dists[, 2L]   # first neighbor is the point itself

  m <- mean(nn)
  sqrt(mean((nn - m)^2)) / m
}

coverage(U)
coverage_fast(U)

C_obs <- coverage_fast(U)

C_null <- replicate(1000, {
  U0 <- matrix(runif(20 * 20), 20, 20)

  # U0 <- matrix(
  #   runif(nrow(U_wrong) * ncol(U_wrong)),
  #   nrow = nrow(U_wrong)
  # )
  coverage_fast(U0)
})

cutoff_95 <- quantile(C_null, .95)
p_value <- (1 + sum(C_null >= C_obs)) / (length(C_null) + 1)

p_value
```

```{r}
test_sim <- function(i) {
  generate_vals <- with(test_mod, rtskewt(n = 1e3, mu = mu, sigmas = sigmas, 
                        skew = skew, nu = nu))

  
}
```

```{r}
discrepancy_result

tibble(x = U[, 1], y = U[, 2]) |> 
  ggplot() +
  geom_point(aes(x = x, y = y), alpha = 0.4) + 
  theme_minimal() +
tibble(x = U_wrong[, 1], y = U_wrong[, 2]) |> 
  ggplot() +
  geom_point(aes(x = x, y = y), alpha = 0.4) + 
  theme_minimal()


n <- nrow(U_matrix)
p <- ncol(U_matrix)

observed <- discrepancy_result$DisL2star

null_rms <- sqrt(
  (2^(-p) - 3^(-p)) / n
)

c(
  observed = observed,
  null_rms = null_rms,
  ratio = observed / null_rms
)

0.0034/null_rms
```


