---
title: "Rosenblatt Derivation"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    code-tools: true
execute: 
  cache: true
---

\newcommand{\coleq}{\mathrel{\mathop:}=}
\newcommand{\ve}[1]{\textbf{#1}}
\newcommand{\field}[1]{\mathbb{#1}}
\newcommand{\defn}[1]{\emph{#1}}
\newcommand{\oneVect}[1]{#1}
\newcommand{\mat}[1]{\textbf{#1}}
\newcommand{\oneMat}[1]{#1}
\newcommand{\tensor}[1]{\mathcal{#1}}
\newcommand{\oneTensor}[1]{#1}
\newcommand{\vs}[1]{\mathcal{#1}}
\newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}

```{r, message = FALSE, echo = FALSE}
library(tidyverse)
library(tensortools)
library(invgamma)
library(patchwork)
library(SHT)
theme_set(theme_minimal())
```

## Rosenblatt transformation

The Rosenblatt transformation is the multivariate extension of the probability integral transformation. It applies the conditional CDF of each element given the elements before it. This removes the modeled dependence sequentially. If the proposed joint distribution is correct, the resulting transformed data are independent Uniform(0, 1) random variables.

Let

$$
X=(X_1,\ldots,X_p)^{T}
$$

be a continuous random vector with joint distribution function $F(x_1, \dots, x_k)$. 

We define the Rosenblatt transformation as follows:

$$
\begin{aligned}
U_1 &=F_{X_1}(X_1),\\
U_2 &=F_{X_2\mid X_1}(X_2\mid X_1),\\
U_3 &=F_{X_3\mid X_1,X_2}(X_3\mid X_1,X_2),\\
&\ \vdots\\
U_p &=F_{X_p\mid X_1,\ldots,X_{p-1}} (X_p\mid X_1,\ldots,X_{p-1}).
\end{aligned}
$$

If the assumed joint distribution is correct and its parameters are known,
then

$$
U_1,\ldots,U_p \stackrel{\mathrm{ind}}{\sim} \text{Uniform}(0,1).
$$

Each random variable is transformed using its distribution conditional on all the random variables that came before it.

## Skewed tensor variate distributions 

The tensor skew-$t$, variance-gamma, inverse-Gaussian, and generalized
hyperbolic distributions implemented in `tensortools` share the representation

$$
\mathcal X = \mathcal M + W\mathcal A+ \sqrt{W}\,\mathcal Z,
$$

where $\mathcal A$ is the skew tensor, $W$ is a positive scalar mixing
variable, and

$$
\mathcal Z\sim
\operatorname{TVN}(\mathbf 0,\Sigma_1,\ldots,\Sigma_D)
$$

is independent of $W$. Conditional on $W=w$,

$$X \sim N(\mathcal{M} + w \mathcal{A}, w [\Sigma_{D} \otimes \dots \Sigma_{1}])$$

## Univariate Skewed-t 

Take $X = \mu + W A + \sigma \sqrt{W} Z, \quad Z \sim N(0, 1)$ where $W > 0$ is a latent mixing variable. Conditional on $W = w, X|W = w \sim N(\mu + wa, w \sigma^2)$. That is, if we knew the value $w$ of the latent mixing variable, $X$ would follow a normal distribution with mean $\mu + w a$ and variance $w \sigma^{2}$. 

In the univariate case, the Rosenblatt transformation is the same as the probability integral transform. 

First, we can obtain the marginal CDF of $X$ by integrating out the dependence on $W$.  
$$
\begin{aligned}
F_{X}(x) &= P(X \le x) \\ 
&= \int_{0}^{\infty} P(X \le x|W = w) f_{W}(w) dw \\
&= \int_{0}^{\infty} f_{X|W}(x|w) f_W(w) dw \\
&= \int_{0}^{\infty} \Phi\left(\frac{x-\mu-wa}{\sigma \sqrt{w}}\right) f_W(w) dw
\end{aligned}
$$ 


Then, we can apply this CDF to the draws. If the model is correct, $U \sim \text{Uniform}(0, 1)$. 

```{r}
#| code-summary: Generate 1e4 univariate skew-t draws

nu <- 4 # generate univariate values
mu <- 1
variance <- 2
skew <- 3

# create univariate draws
univar_skewt <- rtskewt(n = 1e4, mu = mu, sigmas = list(matrix(variance)), 
                        skew = skew, nu = nu)

# define vectorized skewt for ggplot
dtskewt_vec <- function(x, mu, skew, sigmas, nu, log = FALSE) {
  if (length(mu) != 1L || length(skew) != 1L ||
      length(sigmas) != 1L ||
      !all(dim(sigmas[[1]]) == c(1L, 1L))) {
    stop("dtskewt_vec() is only for the univariate case.")
  }

  vapply(
    x,
    function(xi) {
      dtskewt(x = xi, mu = mu, skew = skew, sigmas = sigmas, 
              nu = nu, log = log)
    },
    numeric(1)
  )
}
```


Below, you can see the plot of the skewed-t distribution. The randomly generated draws are plotted with an overlay of the expected distribution. 
```{r, warning = FALSE}
#| code-summary: Plot skewed-t draws with expected distribution

ggplot() +
  geom_histogram(aes(x = univar_skewt, y = after_stat(density)),
                     bins = 200) +
  geom_function(fun = dtskewt_vec, 
                args = list(mu = mu, skew = skew,
                sigmas = list(matrix(variance)), nu = nu),
                color = "red", lty = 2) +
  xlim(quantile(univar_skewt, 0.01), quantile(univar_skewt, 0.99)) + 
  labs(x = "Univariate Skewed-T Draws", y = "Density")
```


We can apply the Rosenblatt transformation to transform the draws from the skewed-t distribution to a uniform. Since these draws are univariate, the Rosenblatt is identical to the Probability Integral Transformation in this case. We apply the CDF to the draws to obtain a uniform. 
```{r}
#| code-summary: Apply Rosenblatt transform to univar bivar skewt

skew_t_cdf <- function(x, mu, variance, skew, nu) {
  shape <- nu/2
  rate <- nu/2

  integrand <- function(w) {
    pnorm((x - mu - w * skew) / sqrt(w * variance)) *  # f(x|w)
    dinvgamma(w, shape = shape, rate = rate) # f(w)
  }

  # integrate out dependence on w
  integrate(integrand, lower = 0, upper = Inf)$value
}

res <- numeric(length(univar_skewt))

for (i in seq_along(univar_skewt)) { # apply to each draw
  res[i] <- skew_t_cdf(x = univar_skewt[i], mu = mu, variance = variance,
                       skew = skew, nu = nu)
}

ggplot() +
  geom_histogram(aes(x = res, y = after_stat(density)), binwidth = 0.05,
                 color = "black", fill = "white") +
  geom_function(fun = dunif, color = "red", lty = 2) +
  labs(x = "Value", y = "Density") 
```

## Multivariate Skewed-t

Now suppose we observe a bivariate skewed-t.

$X = \mu + W \mathbf{a} + \sqrt{W} \mathbf{Z}, \quad Z \sim N_{2}(0, \Sigma)$

where 

$$\Sigma = 
\begin{pmatrix}
\sigma_{1}^{2} & \rho \sigma_{1} \sigma_{2} \\
\rho \sigma_{1} \sigma_{2} & \sigma_{2}^{2}
\end{pmatrix}
$$

Now we need the Rosenblatt to compute $U_{1} = F_{X_{1}}(x_{1})$ and $U_{2} = F_{X_{2}|X_{1}}(x_{2}|x_{1}).$

### Multivariate Skewed-t $X_{1}$ 

First, consider $X_{1}$. Conditional on $W = w, X_{1}|W = w \sim N(\mu_{1} + w a_{1}, w \sigma_{1}^{2})$. This is the same as the univariate case we defined above. 

```{r}
#| code-summary: Generate 1e4 bivariate skew-t draws

nu <- 4 # generate bivariate values
mu <- c(1, 2)
variance <- matrix(data = c(1, 0.5, 0.5, 3), nrow = 2)
skew <- c(0.2, 0.4)

bivar_skewt <- rtskewt(n = 1e4, mu = mu, sigmas = list(variance),
                       skew = skew, nu = nu)
```

### Multivariate Skewed-t $X_{2}$ 

Once we have observed $X_{1} = x_{1}$, we have more information about the mixing variable $W$. Instead of using $f_{W}(w)$, we want to find $f_{W|X_{1}}(w|x_{1})$. 

We can use Bayes' rule to compute 

$$f_{W|X_{1}}(w|x_{1}) \propto f_{X_{1}|W}(x_{1}|w) f_{W}(w)$$

We can then compute $U_{2}$ using the updated CDF of $X_{2}$. 

$$F_{X_{2}|X_{1}}(x_{2}|x_{1}) = \int_{0}^{\infty} F_{X_{2}|X_{1}, W)}(x_{2}|x_{1}, w) f_{W|X_{1}}(w|x_{1}) dw$$

We can apply the CDF to the draws. If the model is correct, they should become Uniform(0, 1). 

```{r}
#| code-summary: Apply Rosenblatt to bivariate skew-t draws

n <- n_draws(bivar_skewt)

L <- t(chol(variance))

b <- drop(forwardsolve(L, skew))

centered <- bivar_skewt - mu

Y <- t(forwardsolve(L, t(centered)))

conditional_u2 <- function(x1, x2, b1, b2, nu) {
  weight <- function(w) {
    dnorm(x1, mean = w * b1, sd = sqrt(w)) *
      dinvgamma(w, shape = nu / 2, rate = nu / 2)
  }

  numerator <- integrate(
    function(w) {
      pnorm(x2, mean = w * b2, sd = sqrt(w)) * weight(w)
    },
    lower = 0,
    upper = Inf
  )$value

  denominator <- integrate(
    weight,
    lower = 0,
    upper = Inf
  )$value

  numerator / denominator
}

u1 <- vapply(
  seq_len(n), 
  function(i) {
    skew_t_cdf(
      x        = bivar_skewt[i][1],
      mu       = mu[1],
      variance = variance[1, 1],
      skew     = skew[1],
      nu       = nu
    )
  },
  numeric(1)
)

# integrate() itself is not vectorized over observations
u2 <- vapply(
  seq_len(n),
  function(i) {
    conditional_u2(
      x1 = Y[i, 1],
      x2 = Y[i, 2],
      b1 = b[1],
      b2 = b[2],
      nu = nu
    )
  },
  numeric(1)
)

res_whitened <- cbind(u1 = u1, u2 = u2)
```

In the plots below, we can see how the Rosenblatt transformation works. The first plot contains the original draws. The second plot is after the Rosenblatt has transformed $X_{1}$ into $U_{1}$, which should now be uniformly distributed along that 1st dimension. The third plot is after the Rosenblatt has transformed $X_{2}$ into $U_{2}$. Now all of the points are uniformly distributed in the space $\mathbb{R}^{2}$.
```{r}
#| code-summary: Plot Rosenblatt applied to $X_{1}$ and $X_{2}$ 

draws_orig <- tibble(x1 = rep(0, 1e4), x2 = rep(0, 1e4))

for(i in 1:1e4) {
  draws_orig$x1[i] <- bivar_skewt[i][1]
  draws_orig$x2[i] <- bivar_skewt[i][2]
}

(draws_orig |> 
  ggplot() +
  geom_point(aes(x = x1, y = x2), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "X1", y = "X2")) +
(draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1, y = x2), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "U1", y = "X2")) +
(draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1, y = u2), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "U1", y = "U2"))
```


## Wrong Parameters

We can intentionally run the Rosenblatt transformation with wrong parameters to see its effect. The true variance used to generate the draws was 

$$\Sigma = 
\begin{pmatrix}
1 & 0.5 \\
0.5 & 3
\end{pmatrix}
$$

Now, we replace $\Sigma_{1, 1}$ with 3. 
```{r}
#| code-summary: Apply Rosenblatt with wrong $\Sigma_{1, 1}$ to bivariate skew-t draws

#true variance: c(1, 0.5, 0.5, 3)
#replace true parameter value
wrong_variance <- matrix(data = c(3, 0.5, 0.5, 3), nrow = 2)

L <- t(chol(wrong_variance))

b <- drop(forwardsolve(L, skew))

centered <- bivar_skewt - mu

Y <- t(forwardsolve(L, t(centered)))

u1_wrong <- vapply(
  seq_len(n), 
  function(i) {
    skew_t_cdf(
      x        = bivar_skewt[i][1],
      mu       = mu[1],
      variance = wrong_variance[1, 1],
      skew     = skew[1],
      nu       = nu
    )
  },
  numeric(1)
)

# integrate() itself is not vectorized over observations
u2_wrong <- vapply(
  seq_len(n),
  function(i) {
    conditional_u2(
      x1 = Y[i, 1],
      x2 = Y[i, 2],
      b1 = b[1],
      b2 = b[2],
      nu = nu
    )
  },
  numeric(1)
)

res_whitened_wrong <- cbind(u1 = u1_wrong, u2 = u2_wrong)
```


Below is the effect of applying the Rosenblatt distribution with the incorrect variance term. The final draws plotted in the $U_{1}$ by $U_{2}$ are not uniform across the space $\mathbb{R}^{2}$.
```{r}
#| code-summary: Plot Rosenblatt with incorrect $\Sigma_{1, 1}$ applied to $X_{1}$ and $X_{2}$ 

(draws_orig |> 
  ggplot() +
  geom_point(aes(x = x1, y = x2), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "X1", y = "X2")) +
(draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1_wrong, y = x2), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "U1", y = "X2")) +
(draws_orig |> 
  ggplot() +
  geom_point(aes(x = u1_wrong, y = u2_wrong), size = 0.6, alpha = 0.25) +
  theme_minimal() +
  labs(x = "U1", y = "U2"))
```


## Formal Diagnostic
For now, assume we have the correct parameter values. The next issue is we need to assess whether the transformed draws are uniform on the unit hypercube. This can be done using an Interpoint Distances test (Yang and Modarres, 2014). 

We define the interpoint distances as 

$$\sum_{i<j}^{n} ||X_{i} - X_{j}||^2$$.

We can define the sample mean of the interpoint distances as 

$$\bar{\delta} = \frac{1}{n \choose 2} \sum_{i<j}^{n} ||X_{i} - X_{j}||^2$$
And the sample variance of the interpoint distances. 

$$\bar{S} = \frac{1}{n \choose 2} \sum_{i<j}^{n} \left(||X_{i} - X_{j}||^{2} - \frac{d}{6} \right)^{2}$$

We want to apply the Central Limit Theorem, so we need to find the variances. 

$$\text{Var}(\hat{\delta}) = \frac{d(2n+3)}{90n(n-1)}$$ and

$$\text{Var}(\bar{S}) = 
\begin{cases}
\frac{1}{n \choose 2} \left(\frac{1}{56700}[989 + 202(n-2)]\right), & \quad d = 2 \\
\frac{1}{n \choose 2} \left(\frac{1}{1050}[37 + 6(n-2)]\right), & \quad d = 3 \\
\frac{1}{n \choose 2} \left(\frac{49 d^{2}}{16200} + \frac{101 d}{37800} + 2(n-2) \left[\frac{d^{2}}{16200} + \frac{29d}{37800}\right]\right), & \quad d \ge 4
\end{cases}$$

Let $\sigma_{\bar{\delta}} = \sqrt{\text{Var}(\bar{\delta})}$ and $\sigma_{\bar{S}} = \sqrt{\text{Var}(\bar{S})}$. 

Then with the Central Limit Theorem for U_processes, we can say that

$$\frac{\bar{\delta} - d/6}{\sigma_{\bar{\delta}}} \xrightarrow{D}  N(0, 1)$$ and

$$\frac{\bar{S} - 7d/180}{\sigma_{\bar{S}}} \xrightarrow{D}  N(0, 1)$$

We define 

$$Q_{1} = \frac{(\bar{\delta} - d/6)^{2}}{\text{Var}(\bar{\delta})}$$ and

$$Q_{2} = \frac{(\bar{S} - 7d/180)^{2}}{\text{Var}(\bar{S})}$$.

Then $Q_{1} \xrightarrow{D} \chi^{2}_{1}$ and $Q_{2} \xrightarrow{D} \chi_{1}^{2}$. 

Since the mean $\bar{\delta}$ and the variance $\bar{S}$ of asymptotically normal variables are independent, we can also define $Q_{3} = Q_{1} + Q_{2} \xrightarrow{D} \chi_{2}^{2}$.

# Test Method

This method has been implemented in the package `SHT` (Statistical Hypothesis Testing Toolbox) as the function `unif.2017YMi`. However, it is inefficient as it constructs the entire $n \times n$ matrix of pairwise distances. The rewritten function below instead takes advantage of vectorized operations to be significantly faster. 
```{r}
#| code-summary: Define an interpoint distances function

interpoint_dist <- function(X) {
  X_array <- if (inherits(X, "tensor")) unclass(X) else X
  X_dims <- dim(X_array)

  if (is.null(X_dims) || length(X_dims) < 2L) {
    stop("X must be a matrix or an array with observations on its first axis.")
  }

  n <- X_dims[1L]
  d <- prod(X_dims[-1L])

  if (n < 2 || d < 2)
    stop("Need at least two observations and two dimensions.")

  X_mat <- matrix(as.numeric(X_array), nrow = n, ncol = d) # matricize data

  pairs <- choose(n, 2) #every possible combination
  fac <- 1 / pairs
  
  values_sq <- rowSums(X_mat^2) #computes x_{i}^2, each obs squared
  obs <- colSums(X_mat) #computes x_{ik}, total of coord across all samples

  # Sum of squared pairwise distances
  sum_pairwise <- n * sum(values_sq) - sum(obs^2)

  # Sum of fourth powers of pairwise distances
  values_four <- sum(values_sq^2)
  v <- drop(crossprod(values_sq, X_mat))

  # Only the squared Frobenius norm is needed.  Use the smaller Gram matrix:
  # ||X'X||_F^2 = ||XX'||_F^2.
  cross_sq <- if (d <= n) {
    cross <- crossprod(X_mat)
    sum(cross * cross)
  } else {
    gram_obs <- tcrossprod(X_mat)
    sum(gram_obs * gram_obs)
  }

  S2 <- (n - 2) * values_four + sum(values_sq)^2 -
        4 * (sum(obs * v) - values_four) +
        2 * (cross_sq - values_four)

  q1_num <- fac * sum_pairwise

  c <- d / 6
  
  q2_num <- fac * (S2 - 2 * c * sum_pairwise + pairs * c^2)

  q1_denom <- d * (2 * n + 3) / (90 * n * (n - 1))

    q2_denom <- if (d == 2) {
    ((989 + 202 * (n - 2)) / 56700) * fac
  } else if (d == 3) {
    ((37 + 6 * (n - 2)) / 1050) * fac
  } else {
    (49 * d^2 / 16200 +
     101 * d / 37800 +
     2 * (n - 2) * (d^2 / 16200 + 29 * d / 37800)) * fac
  }

  Q1 <- (q1_num - d / 6)^2 / q1_denom
  Q2 <- (q2_num - 7 * d / 180)^2 / q2_denom
  Q3 <- Q1 + Q2

  tibble(Q1 = Q1, Q2 = Q2, Q3 = Q3, 
         p_Q1 = pchisq(Q1, df = 1, lower.tail = FALSE),
         p_Q2 = pchisq(Q2, df = 1, lower.tail = FALSE),
         p_Q3 = pchisq(Q3, df = 2, lower.tail = FALSE))
}

bench::mark(unname(SHT::unif.2017YMi(res_whitened, "Q1")$statistic), 
                   interpoint_dist(res_whitened)$Q1)
```


First, we can test that this metric works when we generate truly uniform values. As expected, the rejection rate is about 0.05. However, $Q_{3}$ had a slightly higher rejection rate. $Q_{1}$ and $Q_{2}$ are both asymptotically distributed as a $\chi_{1}^{2}$. $Q_{3}$ is defined as the sum of $Q_{1}$ and $Q_{2}$, so it depends on the joint behavior. This may lead to the finite-sample distribution of $Q_{3}$ to derivate from its asympotitic distribution of $\chi_{2}^{2}$. Due to this behavior, I will use the $p$-values from $Q_{1}$ and $Q_{2}$ as the main diagnostic values for rejection, but will still report $Q_{3}$ in case it is of interest.  

```{r, eval = FALSE}
#| code-summary: Test interpoint distances on truly uniform data

sim_uniform <- function(i) {
  p <- length(mu)

  X <- matrix(
    runif(1e4 * p),
    nrow = 1e4,
    ncol = p
  )

  interpoint_dist(X)
}

res_uniform <- map_dfr(1:1e4, sim_uniform)

saveRDS(res_uniform, "res_uniform_interpoint.rds")
```

```{r, warning = FALSE}
#| code-summary: Report Type I error on truly uniform data

res_uniform <- readRDS("res_uniform_interpoint.rds")

c(Q1_reject = mean(res_uniform$p_Q1 < 0.05),
  Q2_reject = mean(res_uniform$p_Q2 < 0.05),
  Q3_reject = mean(res_uniform$p_Q3 < 0.05))

res_uniform |> 
  ggplot() +
    geom_histogram(aes(x = Q1, y = after_stat(density)), bins = 200) +
    geom_function(fun = dchisq, args = list(df = 1), color = "red", lty = 2) +
    xlim(c(quantile(res_uniform$Q1, 0.05), quantile(res_uniform$Q1, 0.95))) +
    labs(x = "Q1", y = "Density") +
    theme_minimal() +
res_uniform |> 
  ggplot() +
    geom_histogram(aes(x = Q2, y = after_stat(density)), bins = 200) +
    geom_function(fun = dchisq, args = list(df = 1), color = "red", lty = 2) +
    xlim(c(quantile(res_uniform$Q2, 0.05), quantile(res_uniform$Q2, 0.95))) +
    labs(x = "Q2", y = "Density") +
    theme_minimal() +
res_uniform |> 
  ggplot() +
    geom_histogram(aes(x = Q3, y = after_stat(density)), bins = 200) +
    geom_function(fun = dchisq, args = list(df = 2), color = "red", lty = 2) +
    labs(x = "Q3", y = "Density") +
    theme_minimal() 
```


We can run a simulation now where we generate bivariate skewed t values, apply the Rosenblatt transformation with the true parameters, and then compute our test statistic using the interpoint distances. The rejection rate is nominal, with about 5% of samples being rejected. 
```{r, eval = FALSE}
#| code-summary: Test interpoint distances on bivariate skew-t data with true parameters

sim_interdistance <- function(i) {
  new_draws <- rtskewt(n = 1e3, mu = mu, sigmas = list(variance), 
                       skew = skew, nu = nu)


  U <- tensor_rosenblatt(draws = new_draws, model = "skewt",
                        parameters = list(mu = mu, sigmas = list(variance), skew = skew, nu = nu))

  interpoint_dist(unclass(U))
}

library(furrr)
plan(multisession, workers = 8)

res_skewt_params <- future_map_dfr(1:500, sim_interdistance, .progress = TRUE, 
                      .options = furrr_options(seed = TRUE))

saveRDS(res_skewt_params, "res_skewt_params.rds")
```

```{r}
#| code-summary: Report Type I error on bivariate skew-t data with true parameters

res_skewt_params <- readRDS("res_skewt_params.rds")

c(Q1_reject = mean(res_skewt_params$p_Q1 < 0.05),
  Q2_reject = mean(res_skewt_params$p_Q2 < 0.05),
  Q3_reject = mean(res_skewt_params$p_Q3 < 0.05))

ggplot() +
  geom_histogram(aes(x = res_skewt_params$p_Q1), binwidth = 0.05, boundary = 0, color = "black", fill = "white") +
  geom_vline(xintercept = 0.05, color = "red", lty = 2) +
  labs(x = "p-value for Q1", y = "Count") +
ggplot() +
  geom_histogram(aes(x = res_skewt_params$p_Q2), binwidth = 0.05, boundary = 0, color = "black", fill = "white") +
  geom_vline(xintercept = 0.05, color = "red", lty = 2) +
  labs(x = "p-value for Q2", y = "Count") +
  ggplot() +
geom_histogram(aes(x = res_skewt_params$p_Q3), binwidth = 0.05, boundary = 0, color = "black", fill = "white") +
  geom_vline(xintercept = 0.05, color = "red", lty = 2) +
  labs(x = "p-value for Q3", y = "Count") 
```


We can repeat this diagnostic with the purposefully incorrect variance. Now, we expect the test to reject frequently as the specified distribution model is not accurate. This means the Rosenblatt transformation is not able to fully transform the data back into a 2-dimensional uniform on the unit hypercube. As expected, the test rejects every single time as the data is not uniform. 
```{r, eval = FALSE}
#| code-summary: Test interpoint distances on bivariate skew-t data with wrong variance

sim_interdistance_wrong <- function(i) {
  new_draws <- rtskewt(n = 1e3, mu = mu, sigmas = list(variance), 
                       skew = skew, nu = nu)


  U_wrong <- tensor_rosenblatt(draws = new_draws, model = "skewt",
                                parameters = list(mu = mu, sigmas = list(wrong_variance), 
                                skew = skew, nu = nu))

  interpoint_dist(unclass(U_wrong))
}

res_skewt_wrong <- future_map_dfr(1:500, sim_interdistance_wrong, 
                                  .progress = TRUE, .options = furrr_options(seed = TRUE))

saveRDS(res_skewt_wrong, "res_skewt_wrong.rds")                            
```

```{r}
#| code-summary: Report Type I error on bivariate skew-t with wrong variance
res_skewt_wrong <- readRDS("res_skewt_wrong.rds")                            

c(Q1_reject = mean(res_skewt_wrong$p_Q1 < 0.05),
  Q2_reject = mean(res_skewt_wrong$p_Q2 < 0.05),
  Q3_reject = mean(res_skewt_wrong$p_Q3 < 0.05)) 
```


# Uniformity test for each marginal

The interpoint distance assesses whether the joint distribution of the Rosenblatt transformed draws are uniform on the $d$-dimensional unit hypercube. However, we also want to know if they are marginally distributed as Uniform(0, 1) and mutually independent.
```{r}
#| code-summary: Test marginal uniform

test_marginal <- function(draws) {
  d_dims <- length(dim(draws))

  draws_array <- unclass(draws)

  p_marginal <- apply(draws_array, MARGIN = 2:d_dims, 
                      FUN = function(z) ks.test(z, "punif", min = 0, max = 1)$p.value)

  # adjust for multiple comparisons
  p.adjust(p_marginal, method = "BH")                    
}

test_marginal(res_whitened)

test_marginal(res_whitened_wrong)
``` 


## Account for MLEs
When running the interpoint distance test and the marginal tests, we used the true parameters used to simulate the data. However, in practice, we would not know these. Instead, we would use the MLEs which are computed from the EM algorithm. This requires an adjustment to the test statistic. 

The interpoint distance is an order-two U-statistic. U-statistics are statistics that average a function applied to all tuples of a fixed size. It is an order-two statistic since it is averaging the Euclidean distance, which is a square function. 

So the function is defined as 

$$h_{1}(X_{i}, X_{j}) = ||X_{i} - X_{j}||^{2}$$

And then the interpoint distance is 

$$\hat{\delta} = {n \choose  2}^{-1} \sum_{i < j} h_{1}(X_{i}, X_{j})$$

## Step by step process

Let us say that $X_{1}, \dots, X_{n} \sim$ Tensor Variate Skewed T($\mathcal{M}, \mathcal{A}, \bigotimes_{i=1}^{d} \Sigma_{i}, \nu$).

1. We use `tensor_mle` to obtain the MLEs, denoted as $\hat{\theta}$.
2. Compute the information matrix for $\hat{\theta}$.
3. Apply the Rosenblatt transformation to remove the independence structure, transforming the draws to uniform. 
4. Compute the regular interpoint distances with no adjustment yet.
5. Calculate the first-order projections. 
6. Estimate the corrected covariances.
7. From the adjusted test statistic. 



```{r}
#| code-summary: Define adjusted interpoint distances for MLE correction

# Accumulate the score cross-products needed for the MLE correction.
# Only one score chunk is held in memory at a time.
tensor_score_crossproducts <- function(draws, Z, fit,
                                       model = "skewt",
                                       chunk_size = 256L) {
  model <- match.arg(model, c("normal", "skewt", "vargamma",
                              "invgauss", "genhyper"))
  n <- n_draws(draws)
  q <- prod(draw_shape(draws))

  if (!is.matrix(Z) || nrow(Z) != n || ncol(Z) != q) {
    stop("Z must be an n x prod(draw_shape(draws)) matrix.")
  }
  if (length(chunk_size) != 1L || !is.finite(chunk_size) ||
      chunk_size < 1L || chunk_size != as.integer(chunk_size)) {
    stop("chunk_size must be one positive integer.")
  }
  chunk_size <- as.integer(chunk_size)

  # Retain only sums needed for H, G, and I.
  sum_a <- 0
  sum_a2 <- 0
  sum_s <- NULL
  sum_as <- NULL
  sum_ss <- NULL

  for (start in seq.int(1L, n, by = chunk_size)) {
    end <- min(start + chunk_size - 1L, n)
    index <- start:end

    # Scores are calculated only for this chunk.
    score_chunk <- tensor_scores(
      draws = slice_draws(draws, index),
      model = model,
      fit = fit
    )
    score_chunk <- as.matrix(score_chunk)

    # First-order projection for Q1.
    z_chunk <- Z[index, , drop = FALSE]
    a_chunk <- rowSums(z_chunk^2 - z_chunk + 1 / 6)

    if (is.null(sum_s)) {
      p <- ncol(score_chunk)
      sum_s <- numeric(p)
      sum_as <- numeric(p)
      sum_ss <- matrix(0, nrow = p, ncol = p)
    }

    sum_a <- sum_a + sum(a_chunk)
    sum_a2 <- sum_a2 + sum(a_chunk^2)
    sum_s <- sum_s + colSums(score_chunk)
    sum_as <- sum_as + drop(crossprod(a_chunk, score_chunk))
    sum_ss <- sum_ss + crossprod(score_chunk)
  }

  # Center by moment identities without making a second pass.
  mean_a <- sum_a / n
  mean_s <- sum_s / n

  H_hat <- sum_a2 / n - mean_a^2
  G_hat <- sum_as / n - mean_a * mean_s
  I_hat <- sum_ss / n - tcrossprod(mean_s)
  I_hat <- (I_hat + t(I_hat)) / 2

  list(H = H_hat, G = G_hat, I = I_hat)
}

adjusted_q1 <- function(draws, theta_hat,
                        rosenblatt_fun,
                        model = "skewt",
                        chunk_size = 256L,
                        U = NULL,
                        n_draws_fun = n_draws) {
  n <- n_draws_fun(draws)

  # Reuse precomputed Rosenblatt values when available.
  if (is.null(U)) {
    U <- rosenblatt_fun(draws, theta_hat)
  }

  Z <- matrix(unclass(U), nrow = n)
  q <- ncol(Z)

  # Raw Q1 U-statistic: mean squared pairwise distance minus q/6.
  row_sq <- rowSums(Z^2)
  pair_sum <- n * sum(row_sq) - sum(colSums(Z)^2)
  U1 <- pair_sum / choose(n, 2) - q / 6

  moments <- tensor_score_crossproducts(
    draws = draws,
    Z = Z,
    fit = theta_hat,
    model = model,
    chunk_size = chunk_size
  )

  # MLE-adjusted covariance; nu = 2 for pairwise interpoint distances.
  beta <- tryCatch(
    solve(moments$I, moments$G),
    error = function(e) {
      stop("The information matrix is singular or ill-conditioned.")
    }
  )
  Sigma_hat <- 4 * (moments$H - sum(moments$G * beta))

  if (!is.finite(Sigma_hat) || Sigma_hat <= 0) {
    stop("Estimated adjusted variance is not positive.")
  }

  statistic <- n * U1^2 / Sigma_hat

  list(
    U1 = U1,
    variance_adjusted = Sigma_hat,
    statistic = statistic,
    p.value = pchisq(statistic, df = 1, lower.tail = FALSE)
  )
}

rosenblatt_fun <- function(draws, fit) {
  tensor_rosenblatt(
    draws = draws,
    model = "skewt",
    parameters = fit
  )
}

score_fun <- function(draws, fit) {
  tensor_scores(
    draws = draws,
    model = "skewt",
    fit = fit
  )
}
```

```{r}
#| code-summary: Compute Q1 and adjusted Q1 for bivariate skew-t using MLEs

compute_p_mle <- function(i) {
  bivar_skewt <- rtskewt(n = 1e3, mu = mu, sigmas = list(variance),
                       skew = skew, nu = nu)

  theta_hat <- tensor_mle(bivar_skewt, model = "skewt",
                          tol = 1e-6)

  # Rosenblatt transform to uniform
  U_hat <- rosenblatt_fun(bivar_skewt, theta_hat)

  # unadjusted for MLEs
  unadj <- interpoint_dist(unclass(U_hat))

  # adjusted for MLEs
  result <- adjusted_q1(
    draws = bivar_skewt,
    theta_hat = theta_hat,
    #rosenblatt_fun = rosenblatt_fun,
    model = "skewt",
    chunk_size = 256L,
    U = U_hat
  )

  tibble(i = i, q1_unadj = unadj$Q1, p_unadj = unadj$p_Q1,
         q1_adj = result$statistic, p_adj = result$p.value)
}
```

```{r, eval = FALSE}
#| code-summary: Parallelize simulation for speed

library(furrr)

plan(multisession, workers = 10)

res_skewt_MLEs <- future_map_dfr(1:500, compute_p_mle, 
                                 .progress = TRUE, .options = furrr_options(seed = TRUE))
```

```{r}
#| code-summary: Report Type I error on bivariate skew-t with MLEs

res_skewt_MLEs <- readRDS("res_skewt_MLEs.rds")

c(rej_unadj = mean(res_skewt_MLEs$p_unadj < 0.05),
  rej_adj = mean(res_skewt_MLEs$p_adj < 0.05))

res_skewt_MLEs |> 
  pivot_longer(cols = c("p_unadj", "p_adj")) |> 
  ggplot() +
  geom_histogram(aes(x = value), fill = "white", color = "black", binwidth = 0.05) +
  facet_wrap(~name) +
  theme_minimal()

res_skewt_MLEs |> 
  ggplot() +
    geom_histogram(aes(x = q1_unadj, y = after_stat(density)), bins = 200) +
    geom_function(fun = dchisq, args = list(df = 1), color = "red", lty = 2) +
    xlim(c(quantile(res_skewt_MLEs$q1_unadj, 0.05), quantile(res_skewt_MLEs$q1_unadj, 0.95))) +
    labs(x = "Q1_unadjusted", y = "Density") +
    theme_minimal() +
res_skewt_MLEs |> 
  ggplot() +
    geom_histogram(aes(x = q1_adj, y = after_stat(density)), bins = 200) +
    geom_function(fun = dchisq, args = list(df = 1), color = "red", lty = 2) +
    xlim(c(quantile(res_skewt_MLEs$q1_adj, 0.05), quantile(res_skewt_MLEs$q1_adj, 0.95))) +
    labs(x = "Q1_adjusted", y = "Density") +
    theme_minimal() 
```