---
title: "Rosenblatt Derivation"
format: html
---

\newcommand{\coleq}{\mathrel{\mathop:}=}
\newcommand{\ve}[1]{\textbf{#1}}
\newcommand{\field}[1]{\mathbb{#1}}
\newcommand{\defn}[1]{\emph{#1}}
\newcommand{\oneVect}[1]{#1}
\newcommand{\mat}[1]{\textbf{#1}}
\newcommand{\oneMat}[1]{#1}
\newcommand{\tensor}[1]{\mathcal{#1}}
\newcommand{\oneTensor}[1]{#1}
\newcommand{\vs}[1]{\mathcal{#1}}
\newcommand{\mn}[2]{\sideset{_{#1}}{_{#2}}{\mathop{\boldsymbol{\times}}}}

# Rosenblatt transformation for a normal distribution

This section derives the Rosenblatt transformation first for a simple
multivariate normal random vector and then for a tensor-variate normal random
variable. The tensor result is the same normal calculation applied using the
Kronecker-structured tensor covariance.

## Rosenblatt transformation

Let

$$
X=(X_1,\ldots,X_p)^\top
$$

be a continuous random vector with joint distribution function $F(x_1, \dots, x_k)$. 

We define the Rosenblatt transformation as follows:

$$
\begin{aligned}
U_1 &=F_{X_1}(X_1),\\
U_2 &=F_{X_2\mid X_1}(X_2\mid X_1),\\
U_3 &=F_{X_3\mid X_1,X_2}(X_3\mid X_1,X_2),\\
&\ \vdots\\
U_p &=F_{X_p\mid X_1,\ldots,X_{p-1}} (X_p\mid X_1,\ldots,X_{p-1}).
\end{aligned}
$$

If the assumed joint distribution is correct and its parameters are known,
then

$$
U_1,\ldots,U_p \stackrel{\mathrm{ind}}{\sim} \operatorname{Uniform}(0,1).
$$

Each random variable is transformed using its distribution conditional on all the random variables that came before it.

## Bivariate normal example

Suppose

$$
\begin{pmatrix}X_1\\X_2\end{pmatrix} \sim N_2\left[\begin{pmatrix}\mu_1\\\mu_2\end{pmatrix},
\begin{pmatrix}
\sigma_1^2 & \rho \sigma_1 \sigma_2\\
\rho\sigma_1\sigma_2 & \sigma_2^2
\end{pmatrix}
\right].
$$

### Step 2.1: Transform $X_1$

The marginal distribution of the first coordinate is

$$
X_1\sim N(\mu_1,\sigma_1^2).
$$

Therefore,

$$
Z_1=\frac{X_1-\mu_1}{\sigma_1}
$$

is standard normal, and the first Rosenblatt coordinate is

$$
U_1 = F_{X_1}(X_1) =\Phi\left(\frac{X_1-\mu_1}{\sigma_1}\right) =\Phi(Z_1).
$$

### Step 2.2: Transform $X_2$

For a bivariate normal distribution,

$$
X_2\mid X_1=x_1 \sim N\left(\mu_2+\rho\frac{\sigma_2}{\sigma_1}(x_1-\mu_1), \ \sigma_2^ (1-\rho^2) \right).
$$

After subtracting the conditional mean and dividing by the conditional
standard deviation, we obtain

$$
Z_2 = \frac{X_2-\mu_2-\rho(\sigma_2/\sigma_1)(X_1-\mu_1)}{\sigma_2\sqrt{1-\rho^2}}.
$$

Conditional on $X_1$, this quantity is standard normal. Thus the second Rosenblatt transformation is

$$
\begin{aligned}
U_2 &=F_{X_2\mid X_1}(X_2\mid X_1)\\
&=\Phi\left[\frac{X_2-\mu_2-\rho(\sigma_2/\sigma_1)(X_1-\mu_1)}{\sigma_2\sqrt{1-\rho^2}} \right]\\
&=\Phi(Z_2).
\end{aligned}
$$

Under the assumed bivariate normal model, $Z_1$ and $Z_2$ are independent
standard normal variables. Consequently, $U_1$ and $U_2$ are independent
uniform variables.

## 3. Write the same calculation using a Cholesky factor

Now let

$$
X\sim N_p(\mu,\Sigma),
$$

where $\Sigma$ is positive definite. Let

$$
\Sigma=LL^\top
$$

be its lower-triangular Cholesky decomposition. The lower-triangular form is
important because the $j$-th transformed variable depends only on $X_1, \dots, X_j$, matching the sequential conditioning used by the
Rosenblatt transformation.

Define the standardization as 

$$
Z=L^{-1}(X-\mu).
$$
Equivalently,

$$
X=\mu+LZ, \qquad Z\sim N_p(0,I_p).
$$

Because $L$ is lower triangular, the first few coordinates of $X$ satisfy

$$
\begin{aligned}
X_1 &=\mu_1+L_{11}Z_1,\\
X_2 &=\mu_2+L_{21}Z_1+L_{22}Z_2,\\
X_3 &=\mu_3+L_{31}Z_1+L_{32}Z_2+L_{33}Z_3.
\end{aligned}
$$

More generally,

$$
X_j = \mu_j + \sum_{k=1}^{j-1}L_{jk}Z_k + L_{jj}Z_j.
$$

### Step 3.1: Condition on the preceding coordinates

Once $X_1,\ldots,X_{j-1}$ have been observed, the corresponding values
$Z_1,\ldots,Z_{j-1}$ are known from forward substitution. Therefore,

$$
X_j\mid X_1,\ldots,X_{j-1}
\sim
N\left(
\mu_j+\sum_{k=1}^{j-1}L_{jk}Z_k,
L_{jj}^2
\right).
$$

### Step 3.2: Standardize using that conditional distribution

The conditional standardized value is

$$
Z_{j} = \frac{X_j-\mu_j-\sum_{k=1}^{j-1}L_{jk}Z_k }{L_{jj}}.
$$

It follows that the $j$-th Rosenblatt coordinate is simply

$$
\boxed{U_j=\Phi(Z_j)},
\qquad
Z=L^{-1}(X-\mu).
$$

Thus the complete normal Rosenblatt transformation can be summarized as

$$
\boxed{U=\Phi{L^{-1}(X-\mu)}}
$$

where $\Phi$ is applied separately to every coordinate.

This is not just arbitrary whitening. Using the lower-triangular Cholesky factor makes each $Z_j$ exactly the standardized residual from the conditional distribution of $X_j$ given the preceding coordinates.

# Rosenblatt transformation for a tensor-variate normal distribution

## 4. Specify the tensor-normal model

Let $\mathcal X$ be a $D$-way tensor with dimensions
$p_1\times\cdots\times p_D$. Suppose

$$
\mathcal X
\sim
\operatorname{TVN}
(\mathcal M,\Sigma_1,\ldots,\Sigma_D),
$$

where $\mathcal M$ is the mean tensor and $\Sigma_d$ describes covariance
along tensor mode $d$.

Let

$$
x=\operatorname{vec}(\mathcal X),
\qquad
\mu=\operatorname{vec}(\mathcal M).
$$

Using the usual column-major vectorization convention,

$$
x
\sim
N_p(\mu,\Sigma),
\qquad
p=\prod_{d=1}^D p_d,
$$

with

$$
\Sigma
=
\Sigma_D\otimes\cdots\otimes\Sigma_1.
$$

This equation is where the complete tensor covariance structure enters the
Rosenblatt transformation.

## 5. Obtain the tensor-normal Cholesky factor

For each mode, take the lower-triangular Cholesky decomposition

$$
\Sigma_d=L_dL_d^\top.
$$

Define

$$
L=L_D\otimes\cdots\otimes L_1.
$$

Using the Kronecker-product identity

$$
(A\otimes B)(C\otimes D)=(AC)\otimes(BD),
$$

we have

$$
\begin{aligned}
LL^\top
&=(L_D\otimes\cdots\otimes L_1)
  (L_D^\top\otimes\cdots\otimes L_1^\top)\\
&=(L_DL_D^\top)\otimes\cdots\otimes(L_1L_1^\top)\\
&=\Sigma_D\otimes\cdots\otimes\Sigma_1\\
&=\Sigma.
\end{aligned}
$$

Therefore, $L$ is the lower-triangular Cholesky factor of the full vectorized
tensor covariance.

## 6. Standardize the tensor observation

Define

$$
z
=
L^{-1}(x-\mu).
$$

Under the tensor-normal model,

$$
z\sim N_p(0,I_p),
$$

so its coordinates are independent standard normal variables. The tensor-normal
Rosenblatt values are therefore

$$
\boxed{
u_j=\Phi(z_j),
\qquad j=1,\ldots,p.
}
$$

Equivalently, all steps can be written together as

$$
\boxed{
u
=
\Phi\left[
(L_D\otimes\cdots\otimes L_1)^{-1}
\left{
\operatorname{vec}(\mathcal X)-
\operatorname{vec}(\mathcal M)
\right}
\right].
}
$$

## 7. Equivalent mode-by-mode tensor calculation

The same standardization can be performed without explicitly constructing the
large Kronecker product. Let

$$
\mathcal E=\mathcal X-\mathcal M.
$$

Then whiten the residual tensor along every mode:

$$
\mathcal Z
=
\mathcal E
\times_1 L_1^{-1}
\times_2 L_2^{-1}
\cdots
\times_D L_D^{-1},
$$

where $\times_d$ denotes multiplication along mode $d$. With compatible
vectorization and mode-product conventions,

$$
\operatorname{vec}(\mathcal Z)
=
(L_D\otimes\cdots\otimes L_1)^{-1}
\operatorname{vec}(\mathcal E)
=z.
$$

Finally, apply $\Phi$ element by element:

$$
\mathcal U=\Phi(\mathcal Z).
$$

This mode-by-mode form uses all of the tensor covariance matrices while
avoiding construction of a potentially very large full covariance matrix.

## 8. What should be observed under a correct model?

For each tensor observation, the derivation produces

$$
u_1,\ldots,u_p.
$$

If the tensor-normal model and its parameters are correct, then these values
should be

$$
u_1,\ldots,u_p
\stackrel{\mathrm{ind}}{\sim}
\operatorname{Uniform}(0,1).
$$

A diagnostic can therefore examine:

1. **Uniformity:** histograms, uniform Q--Q plots, or a uniformity statistic;
2. **Independence within a tensor:** correlations or other dependence measures
   among the transformed coordinates; and
3. **Independence across observations:** dependence among transformed values
   from different independently sampled tensors.

When $\mathcal M$ and the covariance matrices are replaced by estimates from
the same data, the transformed values are not exactly independent uniforms in
finite samples. A formal goodness-of-fit test therefore needs fitted-parameter
calibration, such as a score-adjusted multiplier calibration or a parametric
bootstrap that repeats both model fitting and transformation.

## 9. Summary of the calculation

For one tensor observation:

1. Subtract the fitted or known mean tensor:
   $\mathcal E=\mathcal X-\mathcal M$.
2. Compute a lower Cholesky factor $L_d$ for every mode covariance
   $\Sigma_d$.
3. Whiten $\mathcal E$ along all modes to obtain $\mathcal Z$.
4. Vectorize $\mathcal Z$ in the same order used by the model.
5. Apply the univariate standard-normal CDF to every entry:
   $u_j=\Phi(z_j)$.
6. Under the correct model with known parameters, the resulting values are
   independent $\operatorname{Uniform}(0,1)$ variables.


# Rosenblatt transformation for skewed tensor-variate distributions

## 10. Normal variance--mean mixture representation

The tensor skew-$t$, variance-gamma, inverse-Gaussian, and generalized
hyperbolic distributions implemented in `tensortools` share the representation

$$
\mathcal X
=
\mathcal M+W\mathcal A+\sqrt{W}\,\mathcal Z,
$$

where $\mathcal A$ is the skew tensor, $W$ is a positive scalar mixing
variable, and

$$
\mathcal Z\sim
\operatorname{TVN}(\mathbf 0,\Sigma_1,\ldots,\Sigma_D)
$$

is independent of $W$. Conditional on $W=w$,

$$
\operatorname{vec}(\mathcal X)\mid W=w
\sim
N_p\left(
\operatorname{vec}(\mathcal M)+w\operatorname{vec}(\mathcal A),
w(\Sigma_D\otimes\cdots\otimes\Sigma_1)
\right).
$$

Thus the distribution is normal when $W$ is known, but it is skewed and
heavy-tailed after $W$ is integrated out. The same $W$ is shared by every
entry of one tensor observation, which also creates dependence among entries.

## 11. Whiten the location and skew tensors

Let $\Sigma_d=L_dL_d^\top$ and apply the same modewise whitening used for the
tensor-normal model:

$$
\mathcal Y
=
(\mathcal X-\mathcal M)
\times_1L_1^{-1}\cdots\times_DL_D^{-1},
$$

and whiten the skew tensor in the same way:

$$
\mathcal B
=
\mathcal A
\times_1L_1^{-1}\cdots\times_DL_D^{-1}.
$$

After vectorizing in R's column-major order, write

$$
y=\operatorname{vec}(\mathcal Y),
\qquad
b=\operatorname{vec}(\mathcal B).
$$

The whitened coordinates satisfy

$$
Y_j\mid W=w\sim N(wb_j,w),
$$

and they are conditionally independent given $W$. They are not marginally
independent because they share the same unobserved mixing variable.

Consequently, applying $\Phi(Y_j)$ is not a valid Rosenblatt transformation.
Instead, each coordinate must use its distribution conditional on all
previous coordinates, with uncertainty about $W$ integrated out.

## 12. Initial GIG mixing distributions

All four mixing distributions can be represented as generalized inverse
Gaussian distributions or their boundary cases. Using

$$
W\sim\operatorname{GIG}(\lambda,\chi,\psi),
$$

with density kernel

$$
f(w)\propto
w^{\lambda-1}
\exp\left\{-\frac12\left(\frac{\chi}{w}+\psi w\right)\right\},
\qquad w>0,
$$

the starting parameters used by `tensor_rosenblatt()` are:

| Tensor distribution | Initial $\lambda_0$ | Initial $\chi_0$ | Initial $\psi_0$ |
|:--|--:|--:|--:|
| Skew-$t$ | $-\nu/2$ | $\nu$ | $0$ |
| Variance-gamma | $\gamma$ | $0$ | $2\gamma$ |
| Inverse-Gaussian | $-1/2$ | $1$ | $\kappa^2$ |
| Generalized hyperbolic | $\lambda$ | $\omega$ | $\omega$ |

The skew-$t$ boundary is an inverse-gamma distribution, while the
variance-gamma boundary is a gamma distribution.

## 13. Sequential posterior update for the mixing variable

Suppose the conditional distribution of $W$ after observing the first $j-1$
whitened coordinates is

$$
W\mid Y_1=y_1,\ldots,Y_{j-1}=y_{j-1}
\sim
\operatorname{GIG}(\lambda_{j-1},\chi_{j-1},\psi_{j-1}).
$$

The likelihood contribution from coordinate $j$ is

$$
f(y_j\mid w)
\propto
w^{-1/2}
\exp\left\{-\frac{(y_j-wb_j)^2}{2w}\right\}.
$$

Combining this likelihood with the GIG density gives another GIG distribution.
After observing $y_j$, its parameters update as

$$
\boxed{
\lambda_j=\lambda_{j-1}-\frac12,
\qquad
\chi_j=\chi_{j-1}+y_j^2,
\qquad
\psi_j=\psi_{j-1}+b_j^2.
}
$$

Equivalently, after the first $j-1$ coordinates,

$$
\lambda_{j-1}=\lambda_0-\frac{j-1}{2},
\quad
\chi_{j-1}=\chi_0+\sum_{k<j}y_k^2,
\quad
\psi_{j-1}=\psi_0+\sum_{k<j}b_k^2.
$$

## 14. The skewed Rosenblatt coordinate

Before updating with $y_j$, calculate its conditional CDF by averaging the
conditional normal CDF over the current posterior distribution of $W$:

$$
\boxed{
U_j
=
\int_0^\infty
\Phi\left(\frac{y_j-wb_j}{\sqrt w}\right)
f_{\operatorname{GIG}}
\left(w;\lambda_{j-1},\chi_{j-1},\psi_{j-1}\right)
\,dw.
}
$$

The calculation for one tensor observation is therefore:

1. Subtract $\mathcal M$ and whiten the observation to obtain $y$.
2. Whiten $\mathcal A$ to obtain $b$.
3. Initialize $(\lambda_0,\chi_0,\psi_0)$ for the selected distribution.
4. For $j=1,\ldots,p$, integrate the conditional normal CDF over the current
   GIG distribution to obtain $U_j$.
5. Update the three GIG parameters using $y_j$ and $b_j$.
6. Continue in the same column-major coordinate order for every observation.

Under the correct model with known parameters, the resulting coordinates are
independent $\operatorname{Uniform}(0,1)$ variables. The package evaluates the
one-dimensional integral numerically on the log-$W$ scale; it never constructs
a high-dimensional skewed tensor CDF.

## 15. Package implementation

For a skew-$t$ model with known parameters, the transformation is obtained by

```r
U <- tensor_rosenblatt(
  draws,
  model = "skewt",
  parameters = list(
    mu = mu,
    sigmas = sigmas,
    skew = skew,
    nu = nu
  )
)
```

For fitted parameters, the MLE can be passed directly:

```r
fit <- tensor_mle(draws, model = "skewt")

U_fitted <- tensor_rosenblatt(
  draws,
  model = "skewt",
  parameters = fit
)
```

These fitted Rosenblatt values are useful for histograms, Q--Q plots, and
dependence diagnostics. Because the same data estimated the parameters, a
formal goodness-of-fit p-value additionally requires fitted-parameter
calibration, as provided by `tensor_gof_scores()` or by a refitted parametric
bootstrap.


```{r}
#| eval: false
#| include: false

library(tensortools)
set.seed(2)

dims <- c(2L, 2L)
mu <- array(0, dim = dims)
sigmas <- list(diag(2), diag(2))
skew <- array(c(0.4, -0.2, 0.3, 0), dim = dims)
nu <- 7

draws <- rtskewt(
  n = 100,
  mu = mu,
  sigmas = sigmas,
  skew = skew,
  nu = nu
)

fit <- tensor_mle(draws, model = "skewt")

U <- tensor_rosenblatt(
  draws,
  model = "skewt",
  parameters = fit
)

hist(unclass(U))

result <- tensor_gof_scores(
  draws,
  model = "skewt",
  fit = fit,
  calibration = "multiplier",
  B = 1999,
  seed = 2026
)

```

## Empirical type-I error under a true skew-$t$ model

This simulation generates observations from one tensor skew-$t$ distribution
and fits that same model in every iteration. The null hypothesis is therefore
true. The rejection proportion estimates the test's type-I error rate and
should be close to the selected significance level in a sufficiently large
simulation.

The current setting uses 100 iterations. At $\alpha=0.05$, approximately five
rejections are expected on average when the test is correctly calibrated,
although the realized count varies. A reliable type-I error study should use
more Monte Carlo iterations, for example at least 500.

```{r skewt-gof-type-one-error}
#| eval: false
#| include: false

type_one_iterations <- 100L
type_one_sample_size <- 100L
type_one_alpha <- 0.05
type_one_B <- 1999L

type_one_dims <- c(2L, 2L)
type_one_mu <- array(0, dim = type_one_dims)
type_one_sigmas <- list(
  matrix(c(1, 0.3, 0.3, 1), nrow = 2),
  toeplitz(c(1, 0.25))
)
type_one_skew <- array(
  c(0.4, -0.2, 0.3, 0),
  dim = type_one_dims
)
type_one_nu <- 7

type_one_results <- vector("list", type_one_iterations)

for (iteration in seq_len(type_one_iterations)) {
  # Giving each iteration its own seed makes individual failures reproducible.
  set.seed(12000 + iteration)

  null_draws <- rtskewt(
    n = type_one_sample_size,
    mu = type_one_mu,
    sigmas = type_one_sigmas,
    skew = type_one_skew,
    nu = type_one_nu
  )

  type_one_results[[iteration]] <- tryCatch(
    {
      null_fit <- tensor_mle(
        null_draws,
        model = "skewt",
        quiet = TRUE
      )

      null_test <- tensor_gof_scores(
        draws = null_draws,
        model = "skewt",
        fit = null_fit,
        marginal_degree = 3L,
        include_interactions = TRUE,
        max_basis = 20L,
        calibration = "multiplier",
        B = type_one_B,
        seed = 22000 + iteration
      )

      data.frame(
        iteration = iteration,
        p_value = null_test$p.value,
        reject = null_test$p.value < type_one_alpha,
        error = NA_character_
      )
    },
    error = function(error) {
      data.frame(
        iteration = iteration,
        p_value = NA_real_,
        reject = NA,
        error = conditionMessage(error)
      )
    }
  )
}

type_one_results <- do.call(rbind, type_one_results)
successful_iterations <- !is.na(type_one_results$p_value)
if (!any(successful_iterations)) {
  stop("All type-I error simulation iterations failed.")
}
estimated_type_one_error <- mean(
  type_one_results$reject[successful_iterations]
)
number_rejected <- sum(
  type_one_results$reject[successful_iterations]
)
type_one_interval <- binom.test(
  number_rejected,
  sum(successful_iterations)
)$conf.int
type_one_mcse_under_nominal <- sqrt(
  type_one_alpha * (1 - type_one_alpha) /
    sum(successful_iterations)
)

type_one_results
data.frame(
  successful_iterations = sum(successful_iterations),
  failed_iterations = sum(!successful_iterations),
  number_rejected = number_rejected,
  nominal_alpha = type_one_alpha,
  estimated_type_one_error = estimated_type_one_error,
  exact_95_percent_lower = type_one_interval[1L],
  exact_95_percent_upper = type_one_interval[2L],
  mcse_under_nominal_alpha = type_one_mcse_under_nominal
)
```
